---
title: "1000G_single_postion_comp"
author: "Andy Beck"
date: "2021-11-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# 1000G Single Position Models - Population Comparisons

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(corrr)
library(tikzDevice)
source("code/single_pos_functions.R")
```

## Introduction

In [another document](1000G_single_position_ALL.Rmd), we evaluated the single position models using singletons aggregated across all five of the 1000 Genomes super-populations. A natural question we can ask is if there is any heterogeneity across the populations. To address this, we perform our analyses again with the singletons, but this time we analyze each super-population separately. Having done this, we assess consistency of the results by checking the correlations of not only the position-level results, but also the contribution to each statistic from the nucleotides at the flanking position under consideration.

## Simple Correlation Analysis

### AT_CG

#### ALL

```{r echo=FALSE}
plot_pos_stat("ALL", "AT_CG", zhu_by_pos, title_text = "Relative Entropy by Relative Position: ", ylab_text = "Relative Entropy")
```

As a reminder, here we plot the single position statistics for the A(T)>C(G) sub-type using singletons aggregated across all 5 super-populations.

#### AFR

```{r echo=FALSE}
plot_pos_stat("AFR", "AT_CG", zhu_by_pos, title_text = "Relative Entropy by Relative Position: ", ylab_text = "Relative Entropy")
```

When we look at the AFR super-population, our results across all the positions looks strikingly similar to what we saw above in the ALL analysis. This might be driven by AFR contributing a larger fraction of the singletons than the other 4 super-populations, so let us now take a look at those.

#### AMR

```{r echo=FALSE}
plot_pos_stat("AMR", "AT_CG", zhu_by_pos, title_text = "Relative Entropy by Relative Position: ", ylab_text = "Relative Entropy")
```

#### EAS

```{r echo=FALSE}
plot_pos_stat("EAS", "AT_CG", zhu_by_pos, title_text = "Relative Entropy by Relative Position: ", ylab_text = "Relative Entropy")
```

#### EUR

```{r echo=FALSE}
plot_pos_stat("EUR", "AT_CG", zhu_by_pos, title_text = "Relative Entropy by Relative Position: ", ylab_text = "Relative Entropy")
```

#### SAS

```{r echo=FALSE}
plot_pos_stat("SAS", "AT_CG", zhu_by_pos, title_text = "Relative Entropy by Relative Position: ", ylab_text = "Relative Entropy")
```

Across all five super-populations we see the same pattern as we look across positions, suggesting that the results are consistent across the five super-populations. The only difference we really see is a shift in the magnitude of the statistic, which is likely driven by differences in the number of singletons across the super-populations.

To quantify our observation that the statistics are consistent across super-populations, let's look at the pairwise correlations of statistics across positions for each pair of super-populations:

```{r results='asis', echo=FALSE}
df_at_cg <- load_all_results_all_sp("AT_CG")

df_wide_at_cg <- 
  df_at_cg %>%
  select(Nuc, rp, pop, s_re) %>%
  pivot_wider(names_from = pop, values_from = s_re)

df_pos_at_cg <- df_wide_at_cg %>%
  group_by(rp) %>%
  summarize(across(AFR:SAS, ~sum(abs(.))))

df_pos_at_cg %>%
  select(-rp) %>%
  correlate()
```

We see in the above pairwise correlation matrix that the position level statistics are highly correlated between super populations which is what we'd like to see: the patterns of influence at single positions seem to be consistent across 1KGP super-populations. However, this is only a surface-level result; the position-level statistics are the sum of nucleotide-specific contributions, and we'd like to assess whether or not these are also consistent across populations.

```{r results='asis', echo=FALSE}
df_cor_at_cg <- df_wide_at_cg %>%
  group_by(rp) %>%
  summarise(across(AMR:SAS, ~cor(., AFR), .names = "AFR_{.col}"),
            across(EAS:SAS, ~cor(., AMR), .names = "AMR_{.col}"),
            across(EUR:SAS, ~cor(., EAS), .names = "EAS_{.col}"),
            across(SAS, ~cor(., EUR), .names = "EUR_{.col}"))

df_cor_at_cg %>%
  select(rp, starts_with("AFR")) %>%
  pivot_longer(-rp, names_to = "pair", values_to = "cor") %>%
  mutate(pair = str_remove(pair, "AFR_")) %>%
  ggplot(aes(x = rp, y = cor, group_by = pair, colour = pair)) +
  geom_point() + 
  geom_line() +
  ylim(c(0,1)) +
  ggtitle("Nucleotide residual correlations", "Reference Population: AFR; subtype: AT_CG") +
  xlab("Relative Position") +
  ylab("Nucleotide Contribution Correlation") +
  labs(colour = "Population")
```

In the above plot we can see that across all positions, the results for the AFR super-population are highly correlated with the other super-populations. The lowest value appears at 1 where the AFR and SAS populations appear to have a lower correlation. But if we look at the nucleotide-level relative entropy residuals at this site we see that the order of the statistics is shared between the two:

```{r}
df_wide_at_cg %>% 
  filter(rp == 1) %>% 
  select(Nuc, AFR, SAS) %>% 
  pivot_longer(AFR:SAS) %>% 
  ggplot(aes(x = Nuc, y = value, color = name, group = name)) + 
  geom_point() +
  geom_line() +
  ggtitle("AT→CG, +1", "AFR and SAS") +
  xlab("Nucleotide")+
  ylab("Signed Relative Entropy Residual")
```


Let's wrap the above analyses into a function and run them on the other subtypes.

```{r}
corr_analysis <- function(subtype, r_start = 1, stat_col = "s_re"){
  df <- load_all_results_all_sp(subtype, r_start = r_start)
  
  df_wide <- df %>%
    select(Nuc, rp, pop, !!sym(stat_col)) %>%
    pivot_wider(names_from = pop, values_from = !!sym(stat_col))
  
  # position-level statistics
  df_pos <- df_wide_at_cg %>%
    group_by(rp) %>%
    summarize(across(AFR:SAS, ~sum(abs(.))))
  
  df_pos %>%
    select(-rp) %>%
    correlate(quiet = TRUE) %>%
    print()
  
  df_cor <- df_wide %>%
    group_by(rp) %>%
    summarise(across(AMR:SAS, ~cor(., AFR), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR), .names = "EUR_{.col}"))
  p <- df_cor %>%
    select(rp, starts_with("AFR")) %>%
    pivot_longer(-rp, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = rp, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", paste0("Reference Population: AFR; subtype: ", subtype)) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
  print(p)
}
```


### AT_GC

```{r}
subtype <- "AT_GC"
corr_analysis(subtype)
```


### AT_TA

```{r}
subtype <- "AT_TA"
corr_analysis(subtype)
```

### GC_AT

```{r}
subtype <- "GC_AT"
corr_analysis(subtype)
```

### cpg_GC_AT

```{r}
subtype <- "cpg_GC_AT"
corr_analysis(subtype, r_start = 2)
```

### GC_CG

```{r}
subtype <- "GC_CG"
corr_analysis(subtype)
```

### cpg_GC_CG

```{r}
subtype <- "cpg_GC_CG"
corr_analysis(subtype, r_start = 2)
```

### GC_TA

```{r}
subtype <- "GC_TA"
corr_analysis(subtype)
```

### cpg_GC_TA

```{r}
subtype <- "cpg_GC_TA"
corr_analysis(subtype, r_start = 2)
```

## Chi Square Approach

Similar to what we did with the single position models, we can also use the chi square goodness of fit test to evaluate, for each sub-type and at each relative position, how similar are the counts between each super-population, and how does this statistic compare to what we observe comparing each super-population to its control sample? We can also do this to compare the BRIDGES data to the 1000G data as well.

Let's first work this out for a sub type at one position, and then move on from there:

```{r}
df_AFR <- load_sp_results("AFR", "AT_CG", -1) %>%
  select(Nuc, singletons) %>%
  mutate(rate = singletons/sum(singletons))

df_AMR <- load_sp_results("AMR", "AT_CG", -1) %>%
  select(Nuc, singletons) %>%
  full_join({df_AFR %>% select(Nuc, rate)}, by = "Nuc") %>%
  mutate(expected = rate * sum(singletons)) %>%
  mutate(chi_sq = (singletons - expected)^2 / expected)

```



## Compare to BRIDGES

Considering the above results suggested that the super-populations are comparable in terms of the single model results, for comparisons with the BRIDGES data set we'll use the ALL "population" for comparisons.

