---
title: "1000G_summary_stats"
author: "Andy Beck"
date: "2021-11-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r include=FALSE, message = FALSE, warning=FALSE}
library(tidyverse)
library(pals)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

# 1000G Summary Statistics

## Introduction

The purpose of this document is to organize statistics regarding the 1000G data set that might be of interest that don't fall into the other documents.

## Singletons per Sample

Below we plot the density of the number of singletons we observe in each sample. First we'll look at all 2,504 individuals across the five super-populations

```{r}
get_singletons_per_subject <- function(singleton_dir){
  awk_cmd <- paste0("awk -F, '{count[$6]++}END{for(key in count)print(key, count[key])}' ",
                    singleton_dir, 
                    "chr*_annotated.csv")
  df <- vroom::vroom(pipe(awk_cmd), col_names = c("ID", "n_singletons"))
  return(df)
}

 # df_ALL <- get_singletons_per_subject("/net/snowwhite/home/beckandy/research/1000G_LSCI/output/singletons/ALL/")
 # write_csv(df_ALL, "/net/snowwhite/home/beckandy/research/1000G_LSCI/output/singletons_per_subject_ALL.csv")

df_ALL <- read_csv("/net/snowwhite/home/beckandy/research/1000G_LSCI/output/singletons_per_subject_ALL.csv",
                      col_types = cols())
df_ALL %>%
  ggplot(aes(x = n_singletons)) +
  geom_density() +
  ggtitle("Singletons per Subject: 1000G",
          paste0("Mean: ", round(mean(df_ALL$n_singletons), 2),
                 ", SD: ", round(sd(df_ALL$n_singletons), 2),
                 ", Min: " , min(df_ALL$n_singletons) ,
                 ", Max: ", max(df_ALL$n_singletons) )
          )+
  xlab("Number of Singletons")
```

Here we can see that while most of the subjects in 1000G have between 10,000 and 30,000 singletons, we see a few outliers in both tails. Let's see what each of the five super-populations looks like:

```{r}
sample_variables <- read_tsv("/net/snowwhite/home/beckandy/research/1000G_LSCI/reference_data/1000G_samples.tsv",
                             col_types = cols())
samples_ALL <- sample_variables  %>% 
  rename(ID = `Sample name`) %>%
  right_join(df_ALL, by = "ID")

samples_AFR <- samples_ALL %>% filter(`Superpopulation code` == "AFR")

samples_AMR <- samples_ALL %>% filter(`Superpopulation code` == "AMR")

samples_EAS <- samples_ALL %>% filter(`Superpopulation code` == "EAS")

samples_EUR <- samples_ALL %>% filter(`Superpopulation code` == "EUR")

samples_SAS <- samples_ALL %>% filter(`Superpopulation code` == "SAS") 
```

```{r}
samples_ALL %>% 
  ggplot(aes(x = n_singletons, fill = `Superpopulation code`)) + 
  geom_density(alpha=0.3) +
  ggtitle("Samples per Subject: 1000G Super-populations")
```

### Singletons per sample within sub-populations

#### AFR

```{r}
samples_AFR %>%
  ggplot(aes(x = n_singletons, fill = `Population code`)) +
  geom_density(alpha = 0.3) +
  ggtitle("Samples per Subject: 1000G AFR") +
  scale_fill_manual(values = as.vector(cols25(n = 7)))
```


Now that we have sample-level information along with the number of singletons, we might also want to look at some more basic summary statistic information (for example: how many subjects do we have within each super-population? average number of singletons per (super)population? etc).

## Samples per Superpopulation

```{r results='asis'}
knitr::kable(table(samples_ALL$`Superpopulation code`))
```

### Samples per Population

#### AFR

```{r results='asis'}
knitr::kable(table(samples_AFR$`Population code`))
```

#### AMR

```{r results='asis'}
knitr::kable(table(samples_AMR$`Population code`))
```

#### EAS

```{r results='asis'}
knitr::kable(table(samples_EAS$`Population code`))
```

#### EUR

```{r results='asis'}
knitr::kable(table(samples_EUR$`Population code`))
```

#### SAS

```{r results='asis'}
knitr::kable(table(samples_SAS$`Population code`))
```
