---
title: "1000G_two_postions_comp"
author: "Andy Beck"
date: "2021-11-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tikzDevice)
library(corrr)
library(reshape2)
source("code/two_pos_new.R")

# latex plot options
width = 4
height = 4
```

## Position-level results

Here in this section of the document we will assess the correlation of position-level results across pairs of populations for all of the mutation subtypes.

### Get position-level statistic for each superpopulation

```{r}
get_pos <- function(subtype, r_start=1, data_dir = "/net/snowwhite/home/beckandy/research/1000G_LSCI/output/all_count_2_pos/"){
  final <- stat_pair_all("AFR", subtype, zhu_pair_re, r_start = r_start, data_dir = data_dir )
  colnames(final)[3] <- "AFR"
  for(pop in c("AMR", "EAS", "EUR", "SAS")){
    df <- stat_pair_all(pop, subtype, zhu_pair_re, r_start = r_start, data_dir = data_dir)
    colnames(df)[3] <- pop
    final <- full_join(final, df, by = c("p1", "p2"))
  }
  return(final)
}
```

### AT_CG

```{r}
st <- "AT_CG"

simple_cor <- function(st, r_start = 1){
  df <- get_pos(st, r_start = r_start)
  
  cor_mat <- cor({df %>% select(-p1, -p2)})
  
  if(!str_starts(st, "cpg")){
    subtype2 <- str_replace(st, "_", " → ")
  } else {
    subtype <- str_sub(st, 5)
    subtype2 <- paste0("(cpg) ", str_replace(subtype, "_", " → "))
  }
  
  p <- cor({df %>% select(-p1, -p2)}) %>% 
    melt %>%
    ggplot(aes(x=Var1, y=Var2, fill=value)) +
    geom_tile() +
    xlab("Population") + ylab("Population") + 
    ggtitle("Correlation of Position-Level Results", paste0("Subtype: ", subtype2)) +
    scale_fill_distiller(palette = "Reds", direction = 1, limits = c(0,1))
  return(list(p = p, cor_mat = cor_mat))
}
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### AT_GC

```{r}
st <- "AT_GC"
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### AT_TA

```{r}
st <- "AT_TA"
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### GC_AT

```{r}
st <- "GC_AT"
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### cpg_GC_AT

```{r}
st <- "cpg_GC_AT"
obj <- simple_cor(st, r_start = 2)

obj$cor_mat %>% knitr::kable()
obj$p
```

### GC_TA

```{r}
st <- "GC_TA"
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### cpg_GC_TA

```{r}
st <- "cpg_GC_TA"
obj <- simple_cor(st, r_start = 2)

obj$cor_mat %>% knitr::kable()
obj$p
```

### GC_CG

```{r}
st <- "GC_CG"
obj <-  simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### cpg_GC_CG

```{r}
st <- "cpg_GC_CG"
obj <- simple_cor(st, r_start = 2)

obj$cor_mat %>% knitr::kable()
obj$p
```


## Do populations share location of highest RE?

```{r}
get_table_re <- function(pop){
  
  final <- data.frame(st = character(),
                      p1 = numeric(),
                      p2 = numeric(),
                      statistic = numeric())
  
  for(st in c("AT_CG", "AT_GC", "AT_TA", "GC_AT", "GC_TA", "GC_CG")){
    df <- stat_pair_all(pop, st, zhu_pair_re) %>%
      arrange(desc(statistic))
    p1 <- df$p1[1]
    p2 <- df$p2[1]
    statistic <- df$statistic[1]
    final <- bind_rows(final,
                       data.frame(st = st,
                                  p1 = p1,
                                  p2 = p2,
                                  statistic = statistic))
  }
  for(st in c("cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG")){
    df <- stat_pair_all("ALL", st, zhu_pair_re, r_start = 2) %>%
      arrange(desc(statistic))
    p1 <- df$p1[1]
    p2 <- df$p2[1]
    statistic <- df$statistic[1]
    final <- bind_rows(final,
                       data.frame(st = st,
                                  p1 = p1,
                                  p2 = p2,
                                  statistic = statistic))
  }
  
  final$pop <- pop
  return(final)
}
```
### ALL

```{r}
pop <- "ALL"
get_table_re(pop) %>%
  knitr::kable()
```

### AFR

```{r}
pop <- "AFR"
get_table_re(pop) %>%
  knitr::kable()
```

### AMR

```{r}
pop <- "AMR"
get_table_re(pop) %>%
  knitr::kable()
```

### EAS

```{r}
pop <- "EAS"
get_table_re(pop) %>%
  knitr::kable()
```

### EUR

```{r}
pop <- "EUR"
get_table_re(pop) %>%
  knitr::kable()
```

### SAS

```{r}
pop <- "SAS"
get_table_re(pop) %>%
  knitr::kable()
```

## GC_AT Differences

```{r}
stat_pair_all("ALL", "GC_AT", zhu_pair_re) %>%
  arrange(desc(statistic)) %>%
  head(4) %>%
  knitr::kable()

stat_pair_all("AFR", "GC_AT", zhu_pair_re) %>%
  arrange(desc(statistic)) %>%
  head(4) %>%
  knitr::kable()

stat_pair_all("AMR", "GC_AT", zhu_pair_re) %>%
  arrange(desc(statistic)) %>%
  head(4) %>%
  knitr::kable()

stat_pair_all("EAS", "GC_AT", zhu_pair_re) %>%
  arrange(desc(statistic)) %>%
  head(4) %>%
  knitr::kable()

stat_pair_all("EUR", "GC_AT", zhu_pair_re) %>%
  arrange(desc(statistic)) %>%
  head(4) %>%
  knitr::kable()

stat_pair_all("SAS", "GC_AT", zhu_pair_re) %>%
  arrange(desc(statistic)) %>%
  head(4) %>%
  knitr::kable()
```

In this example, while not all populations share the positions coordinates of the highest RE value, we see that (-2, -1) and (-1, 1) are in the top two for all populations.

## AT_CG example (EAS)

For this subtype the only super population to not have the highest observed RE value at (1,2) was EAS (-2, -1)

```{r}
stat_pair_all("ALL", "AT_CG", zhu_pair_re) %>%
  arrange(desc(statistic)) %>%
  head(4) %>%
  knitr::kable()

stat_pair_all("EAS", "AT_CG", zhu_pair_re) %>%
  arrange(desc(statistic)) %>%
  head(4) %>%
  knitr::kable()

stat_pair_all("AFR", "AT_CG", zhu_pair_re) %>%
  arrange(desc(statistic)) %>%
  head(4) %>%
  knitr::kable()

stat_pair_all("AMR", "AT_CG", zhu_pair_re) %>%
  arrange(desc(statistic)) %>%
  head(4) %>%
  knitr::kable()

stat_pair_all("EUR", "AT_CG", zhu_pair_re) %>%
  arrange(desc(statistic)) %>%
  head(4) %>%
  knitr::kable()

stat_pair_all("SAS", "AT_CG", zhu_pair_re) %>%
  arrange(desc(statistic)) %>%
  head(4) %>%
  knitr::kable()
```

Again, we see that across all populations the top four statistics are at the same pairs of positions, with only the order slightly adjusted.
