---
title: "1000G_single_postion_comp"
author: "Andy Beck"
date: "2021-11-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# 1000G Single Position Models - Population Comparisons

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(corrr)
library(tikzDevice)
source("code/single_pos_functions.R")
```

## Introduction

In [another document](1000G_single_position_ALL.Rmd), we evaluated the single position models using singletons aggregated across all five of the 1000 Genomes super-populations. A natural question we can ask is if there is any heterogeneity across the populations. To address this, we perform our analyses again with the singletons, but this time we analyze each super-population separately. Having done this, we assess consistency of the results by checking the correlations of not only the position-level results, but also the contribution to each statistic from the nucleotides at the flanking position under consideration.

## Simple Correlation Analysis

### AT_CG

#### ALL

```{r echo=FALSE}
plot_pos_stat("AT_CG", "ALL")
```

As a reminder, here we plot the single position statistics for the A(T)>C(G) sub-type using singletons aggregated across all 5 super-populations.

#### AFR

```{r echo=FALSE}
plot_pos_stat("AT_CG", "AFR")
```

When we look at the AFR super-population, our results across all the positions looks strikingly similar to what we saw above in the ALL analysis. This might be driven by AFR contributing a larger fraction of the singletons than the other 4 super-populations, so let us now take a look at those.

#### AMR

```{r echo=FALSE}
plot_pos_stat("AT_CG", "AMR")
```

#### EAS

```{r echo=FALSE}
plot_pos_stat("AT_CG", "EAS")
```

#### EUR

```{r echo=FALSE}
plot_pos_stat("AT_CG", "EUR")
```

#### SAS

```{r echo=FALSE}
plot_pos_stat("AT_CG", "SAS")
```

Across all five super-populations we see the same pattern as we look across positions, suggesting that the results are consistent across the five super-populations. The only difference we really see is a shift in the magnitude of the statistic, which is likely driven by differences in the number of singletons across the super-populations.

To quantify our observation that the statistics are consistent across super-populations, let's look at the pairwise correlations of statistics across positions for each pair of super-populations:

```{r results='asis', echo=FALSE}
df_corr_at_cg <- load_all_results_all_sp("AT_CG")
corr_mat_at_cg <- df_corr_at_cg %>%
  select(rp, starts_with("chi_sq_ct")) %>%
  group_by(rp) %>%
  summarize_all(sum) %>%
  ungroup() %>% select(-rp) %>%
  correlate()

corr_mat_at_cg %>% fashion() %>% knitr::kable()
```

We see here that the statistics at each position are highly correlated across the super-populations. We can also ask if the "residuals" (contribution of each nucleotide to each position's statistic) is also correlated:

```{r results='asis', echo=FALSE}
df_corr_at_cg %>%
  select(starts_with("chi_sq_ct")) %>%
  correlate() %>%
  fashion() %>%
  knitr::kable(format = "latex")
```
and it appears that they are for this sub-type. Let's now take a look at the other sub-types

### AT_GC

#### Position-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr <- load_all_results_all_sp("AT_GC")
df_corr %>%
  select(rp, starts_with("chi_sq_ct")) %>%
  group_by(rp) %>%
  summarize_all(sum) %>%
  ungroup() %>% select(-rp) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

#### Nucleotide-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr %>%
  select(starts_with("chi_sq_ct")) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

### AT_TA

#### Position-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr <- load_all_results_all_sp("AT_TA")
df_corr %>%
  select(rp, starts_with("chi_sq_ct")) %>%
  group_by(rp) %>%
  summarize_all(sum) %>%
  ungroup() %>% select(-rp) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

#### Nucleotide-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr %>%
  select(starts_with("chi_sq_ct")) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

### GC_AT

#### Position-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr <- load_all_results_all_sp("GC_AT")
df_corr %>%
  select(rp, starts_with("chi_sq_ct")) %>%
  group_by(rp) %>%
  summarize_all(sum) %>%
  ungroup() %>% select(-rp) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

#### Nucleotide-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr %>%
  select(starts_with("chi_sq_ct")) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

### cpg_GC_AT

#### Position-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr <- load_all_results_all_sp("cpg_GC_AT", r_start = 2)
df_corr %>%
  select(rp, starts_with("chi_sq_ct")) %>%
  group_by(rp) %>%
  summarize_all(sum) %>%
  ungroup() %>% select(-rp) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

#### Nucleotide-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr %>%
  select(starts_with("chi_sq_ct")) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

### GC_CG

#### Position-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr <- load_all_results_all_sp("GC_CG")
df_corr %>%
  select(rp, starts_with("chi_sq_ct")) %>%
  group_by(rp) %>%
  summarize_all(sum) %>%
  ungroup() %>% select(-rp) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

#### Nucleotide-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr %>%
  select(starts_with("chi_sq_ct")) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

### cpg_GC_CG

#### Position-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr <- load_all_results_all_sp("cpg_GC_CG", r_start = 2)
df_corr %>%
  select(rp, starts_with("chi_sq_ct")) %>%
  group_by(rp) %>%
  summarize_all(sum) %>%
  ungroup() %>% select(-rp) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

#### Nucleotide-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr %>%
  select(starts_with("chi_sq_ct")) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

### GC_TA

#### Position-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr <- load_all_results_all_sp("GC_TA")
df_corr %>%
  select(rp, starts_with("chi_sq_ct")) %>%
  group_by(rp) %>%
  summarize_all(sum) %>%
  ungroup() %>% select(-rp) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

#### Nucleotide-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr %>%
  select(starts_with("chi_sq_ct")) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

### cpg_GC_TA

#### Position-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr <- load_all_results_all_sp("cpg_GC_TA", r_start = 2)
df_corr %>%
  select(rp, starts_with("chi_sq_ct")) %>%
  group_by(rp) %>%
  summarize_all(sum) %>%
  ungroup() %>% select(-rp) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

#### Nucleotide-level

```{r results='asis', echo=FALSE, message=FALSE}
df_corr %>%
  select(starts_with("chi_sq_ct")) %>%
  correlate() %>% 
  fashion() %>% 
  knitr::kable()
```

## Chi Square Approach

Similar to what we did with the single position models, we can also use the chi square goodness of fit test to evaluate, for each sub-type and at each relative position, how similar are the counts between each super-population, and how does this statistic compare to what we observe comparing each super-population to its control sample? We can also do this to compare the BRIDGES data to the 1000G data as well.

```{r}
load_singletons_2pop <- function(subtype, sp1, sp2, r_start = 1){
  df1 <- load_all_results(subtype, sp1, r_start) %>%
    filter(singletons > 0) %>%
    select(rp, Nuc, singletons, chi_sq_ct) %>%
    rename_with(.fn = ~paste0(., ".", sp1), .cols = c(starts_with("chi"),starts_with("singletons") ))
  df2 <- load_all_results(subtype, sp2, r_start) %>%
    filter(singletons > 0) %>%
    select(rp, Nuc, singletons, chi_sq_ct) %>%
    rename_with(.fn = ~paste0(., ".", sp2),  .cols = c(starts_with("chi"),starts_with("singletons") ))
  
  df <- full_join(df1, df2, by = c("rp", "Nuc"))
  return(df)
}

chi_sq_2pop <- function(subtype, sp1, sp2, r_start = 1){
  df <- load_singletons_2pop(subtype, sp1, sp2, r_start)
  chi_vec1 <- c()
  chi_vec2 <- c()
  for(i in c(-10:-1, r_start:10)){
    df2 <- df %>% 
      filter(rp == i)
    vec1 <- df2 %>% select(starts_with("singletons") & contains(sp1)) %>% pull(1)
    vec2 <- df2 %>% select(starts_with("singletons") & contains(sp2)) %>% pull(1)
    chi_obj1 <- chisq.test(vec1, p = vec2 / sum(vec2))
    chi_obj2 <- chisq.test(vec2, p = vec1 / sum(vec1))
    chi_vec1 <- c(chi_vec1, unname(chi_obj1$statistic))
    chi_vec2 <- c(chi_vec2, unname(chi_obj2$statistic))
  }
  control_df <- df %>% 
    group_by(rp) %>% 
    summarize_at(vars(starts_with("chi")), sum)
  final <- data.frame(rp = c(-10:-1, r_start:10))
  final[,paste0(sp1, "-", sp2,"")] <- chi_vec1
  final[,paste0(sp2, "-", sp1,"")] <- chi_vec2
  final[,paste0("Control - ", sp1)] <- control_df[,2]
  final[,paste0("Control - ", sp2)] <- control_df[,3]
  return(final)
}
```

For an illustrative example, let's take a look at the A(T)>C(G) sub-type in the EUR and AFR super-populations. Here we'll use the frequencies observed in EUR to fit the AFR counts, and compare these chi-square statistics to those that we obtain when we compare the EUR and AFR super-populations to their corresponding control distributions:

```{r results='asis'}
chi_sq_2pop("AT_CG", "AFR", "EUR") %>%
  knitr::kable()
```

Here we can see that the statistics when we use EUR proportions to fit AFR and vice versa are much smaller that what we observe using the controls to fit either population's counts. We can also plot these to make the exact same conclusion:

```{r echo=FALSE}
chi_sq_2pop_plot <- function(subtype, sp1, sp2, r_start = 1){
  df <- chi_sq_2pop(subtype, sp1, sp2, r_start) %>%
    gather(type, statistic, -rp)
  
  p <- df %>%
    ggplot(aes(x = rp, y = statistic, colour = type)) + 
    geom_point() +
    geom_line() +
    ggtitle(paste0("Single Position Chi Square Statistic: ", str_replace_all(subtype, "_", "-")),
            paste0("Pop 1: ", sp1, "; Pop 2: ", sp2)) +
    xlab("Relative Position") + 
    ylab("Statistic") +
    labs(fill="Statistic Type")
  return(p)
}
```

```{r message=FALSE}
p <- chi_sq_2pop_plot("AT_CG", "AFR", "EUR")

tikz("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/AFR_EUR_AT_CG.tex",
     width = 7, height = 4)
p
dev.off()

p
```

### AT_CG

```{r message=FALSE}
st <- "AT_CG"
for(sp in c("AMR", "EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AFR", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AFR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AMR", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AMR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "EAS", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
p <- chi_sq_2pop_plot(st, "EUR", "SAS")
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_","SAS","_",st,".tex"))
p
dev.off()
print(p)
```

### AT_GC

```{r message=FALSE}
st <- "AT_GC"
for(sp in c("AMR", "EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AFR", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AFR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AMR", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AMR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "EAS", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
p <- chi_sq_2pop_plot(st, "EUR", "SAS")
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_","SAS","_",st,".tex"))
p
dev.off()
print(p)
```

### AT_TA

```{r echo=FALSE, message=FALSE}
st <- "AT_TA"
for(sp in c("AMR", "EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AFR", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AFR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AMR", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AMR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "EAS", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
p <- chi_sq_2pop_plot(st, "EUR", "SAS")
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_","SAS","_",st,".tex"))
p
dev.off()
print(p)
```

### GC_AT

```{r echo=FALSE, message=FALSE}
st <- "GC_AT"
for(sp in c("AMR", "EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AFR", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AFR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AMR", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AMR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "EAS", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
p <- chi_sq_2pop_plot(st, "EUR", "SAS")
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_","SAS","_",st,".tex"))
p
dev.off()
print(p)
```

### cpg_GC_AT

```{r echo=FALSE, message=FALSE}
st <- "cpg_GC_AT"
for(sp in c("AMR", "EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AFR", sp, r_start = 2)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AFR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AMR", sp, r_start = 2)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AMR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "EAS", sp, r_start = 2)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
p <- chi_sq_2pop_plot(st, "EUR", "SAS", r_start = 2)
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_","SAS","_",st,".tex"))
p
dev.off()
print(p)
```

### GC_TA

```{r echo=FALSE, message=FALSE}
st <- "GC_TA"
for(sp in c("AMR", "EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AFR", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AFR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AMR", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AMR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "EAS", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
p <- chi_sq_2pop_plot(st, "EUR", "SAS")
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_","SAS","_",st,".tex"))
p
dev.off()
print(p)
```

### cpg_GC_TA

```{r echo=FALSE, message=FALSE}
st <- "cpg_GC_TA"
for(sp in c("AMR", "EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AFR", sp, r_start = 2)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AFR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AMR", sp, r_start = 2)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AMR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "EAS", sp, r_start = 2)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
p <- chi_sq_2pop_plot(st, "EUR", "SAS", r_start = 2)
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_","SAS","_",st,".tex"))
p
dev.off()
print(p)
```

### GC_CG

```{r echo=FALSE, message=FALSE}
st <- "GC_CG"
for(sp in c("AMR", "EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AFR", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AFR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AMR", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AMR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "EAS", sp)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
p <- chi_sq_2pop_plot(st, "EUR", "SAS")
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_","SAS","_",st,".tex"))
p
dev.off()
print(p)
```

### cpg_GC_CG

```{r echo=FALSE, message=FALSE}
st <- "cpg_GC_CG"
for(sp in c("AMR", "EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AFR", sp, r_start = 2)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AFR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EAS", "EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "AMR", sp, r_start = 2)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "AMR","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
for(sp in c("EUR", "SAS")){
  p <- chi_sq_2pop_plot(st, "EAS", sp, r_start = 2)
  tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_",sp,"_",st,".tex"),
     width = 7, height = 4)
  p
  dev.off()
  print(p)
}
p <- chi_sq_2pop_plot(st, "EUR", "SAS", r_start = 2)
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "EAS","_","SAS","_",st,".tex"))
p
dev.off()
print(p)
```

## Compare to BRIDGES

Considering the above results suggested that the super-populations are comparable in terms of the single model results, for comparisons with the BRIDGES data set we'll use the ALL "population" for comparisons.

```{r echo=FALSE}
statistic_by_position_all <- function(subtype, sp, r_start = 1){
  df_1000G <- statistic_by_position(subtype, sp, r_start) %>%
    mutate(type = paste0(type, "_1000G"))
  df_bridges <- statistic_by_position_bridges(subtype, r_start) %>%
    mutate(type = paste0(type, "_BRIDGES"))
  df <- bind_rows(df_1000G, df_bridges)
  return(df)
}

statistic_by_position_all_plot <- function(subtype, sp, r_start = 1){
  df <- statistic_by_position_all(subtype, sp, r_start)
  p <- df %>%
    filter(str_starts(type, "Control")) %>%
    mutate(type = str_remove(type, "Control_")) %>%
    ggplot(aes(x = pos, y= chi_sq, colour = type)) +
    geom_line() + geom_point() +
    xlab("Relative Position") +
    ylab("Statistic") +
    labs(colour = "Data Set") +
    ggtitle(paste0("BRIDGES and 1000G: ", subtype))
  return(p)
}
```

### AT_CG

```{r}
st <- "AT_CG"

p <- statistic_by_position_all_plot(st, "ALL")
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "BRIDGES","_","1KGP","_",st,".tex"))
p
dev.off()
print(p)
```

### AT_GC

```{r}
st <- "AT_GC"

p <- statistic_by_position_all_plot(st, "ALL")
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "BRIDGES","_","1KGP","_",st,".tex"))
p
dev.off()
print(p)
```

### AT_TA

```{r}
st <- "AT_TA"

p <- statistic_by_position_all_plot(st, "ALL")
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "BRIDGES","_","1KGP","_",st,".tex"))
p
dev.off()
print(p)
```

### GC_AT

```{r}
st <- "GC_AT"

p <- statistic_by_position_all_plot(st, "ALL")
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "BRIDGES","_","1KGP","_",st,".tex"))
p
dev.off()
print(p)
```

### cpg_GC_AT

```{r}
st <- "cpg_GC_AT"

p <- statistic_by_position_all_plot(st, "ALL", r_start = 2)
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "BRIDGES","_","1KGP","_",st,".tex"))
p
dev.off()
print(p)
```

### GC_TA

```{r}
st <- "GC_TA"

p <- statistic_by_position_all_plot(st, "ALL")
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "BRIDGES","_","1KGP","_",st,".tex"))
p
dev.off()
print(p)
```

### cpg_GC_TA

```{r}
st <- "cpg_GC_TA"

p <- statistic_by_position_all_plot(st, "ALL", r_start = 2)
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "BRIDGES","_","1KGP","_",st,".tex"))
p
dev.off()
print(p)
```

### GC_CG

```{r}
st <- "GC_CG"

p <- statistic_by_position_all_plot(st, "ALL")
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "BRIDGES","_","1KGP","_",st,".tex"))
p
dev.off()
print(p)
```

### cpg_GC_CG

```{r}
st <- "cpg_GC_CG"

p <- statistic_by_position_all_plot(st, "ALL", r_start = 2)
tikz(paste0("/net/snowwhite/home/beckandy/research/documentation/LSCI/output/tikz/single_pos/",
              "BRIDGES","_","1KGP","_",st,".tex"))
p
dev.off()
print(p)
```

We see that for all the sub-types the BRIDGES results are highly correlated with the 1000G results. But for the paper I'd like to have a table of correlation values.

```{r}
load_singletons_b1kgp <- function(subtype, r_start = 2){
  df_bridges <- load_all_results_bridges(subtype, r_start) %>%
    select(Nuc, rp, chi_sq_ct) %>%
    rename(BRIDGES = chi_sq_ct)
  df_g <- load_all_results(subtype, "ALL", r_start) %>%
    select(Nuc, rp, chi_sq_ct) %>%
    rename(`1KGP` = chi_sq_ct)
  df <- full_join(df_bridges, df_g, by = c("Nuc", "rp"))
  return(df)
}
```

For all sub types we pull the nucleotide-level statistics for both data sets and get the correlations of both the position-level results and the nucleotide-level results:

```{r}
final <- data.frame(Nuc = character(),
                    rp = numeric(),
                    BRIDGES = numeric(),
                    blank = numeric(),
                    cat = character()) %>%
  rename(`1KGP` = blank)

for(st in c("AT_CG", "AT_GC", "AT_TA", "GC_AT", "GC_TA", "GC_CG")){
  df <- load_singletons_b1kgp(st) %>%
    mutate(cat = st)
  final <- bind_rows(final, df)
}

for(st in c("cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG")){
  df <- load_singletons_b1kgp(st, r_start = 2) %>%
    mutate(cat = st)
  final <- bind_rows(final, df)
}
```

Let's first get the correlation across nucleotide-level residuals:

```{r, results='asis'}
final %>%
  group_by(cat) %>%
  summarize(cor = cor(`1KGP`, BRIDGES)) %>%
  knitr::kable(format = "latex")
```

And now let's get the correlations of the position-level statistics:

```{r, results='asis'}
final %>%
  group_by(cat, rp) %>%
  summarize(tgp = sum(`1KGP`), br = sum(BRIDGES)) %>%
  group_by(cat) %>%
  summarize(correlation = cor(tgp, br)) %>%
  knitr::kable(format = "latex")
```

