---
title: "BRIDGES_summary_stats"
author: "Andy Beck"
date: "2021-11-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r include=FALSE, message = FALSE, warning=FALSE}
library(tidyverse)
library(tikzDevice)
```

# BRIDGES Summary Statistics

## Introduction

The purpose of this document is to organize statistics regarding the BRIDGES data set that might be of interest that don't fall into the other documents.

## Singletons per Sample

Below we plot the density of the number of singletons we observe in each sample.

```{r}
get_singletons_per_subject <- function(singleton_dir){
  awk_cmd <- paste0("awk -F, '{count[$6]++}END{for(key in count)print(key, count[key])}' ",
                    singleton_dir, 
                    "chr*_annotated.csv")
  df <- vroom::vroom(pipe(awk_cmd), col_names = c("ID", "n_singletons"))
  return(df)
}

# df_BRIDGES <- get_singletons_per_subject("/net/snowwhite/home/beckandy/research/BRIDGES_redo/output/singletons/")
# write_csv(df_BRIDGES, "/net/snowwhite/home/beckandy/research/BRIDGES_redo/output/singletons_per_subject.csv")

df_BRIDGES <- read_csv("/net/snowwhite/home/beckandy/research/BRIDGES_redo/output/singletons_per_subject.csv",
                       col_types = cols())
p <- df_BRIDGES %>%
  ggplot(aes(x = n_singletons)) + 
  geom_density() +
  ggtitle("Singletons per Subject: BRIDGES",
          paste0("Mean: ", round(mean(df_BRIDGES$n_singletons), 2),
                 ", SD: ", round(sd(df_BRIDGES$n_singletons), 2),
                 ", Min: " , min(df_BRIDGES$n_singletons) ,
                 ", Max: ", max(df_BRIDGES$n_singletons) ) 
          )+
  xlab("Number of Singletons")

p

tikz("output/tikz/singletons_per_sub_BRIDGES.tex")
print(p)
dev.off()
```

Here, we can see that on average we see around 10,000 singletons per sample in the BRIDGES data set, although we do have outliers in both directions.

**Question: Are these outlier samples different in other ways? E.g., 3-mer mutation spectra, spatial distribution of nucleotides genome-wide, etc?**

## Singletons per subtype

```{r}
data_dir <- "/net/snowwhite/home/beckandy/research/BRIDGES_redo/output/singletons/"
subtypes <- c("AT_CG","AT_GC","AT_TA", "GC_AT", "GC_TA", "GC_CG", "cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG")
final <- data.frame(subtype = character(), n = numeric())
for(st in subtypes){
  n <- system2("wc", args = c("-l", paste0(data_dir, st, ".txt"), " | awk '{print $1}'"), stdout = TRUE) %>% as.numeric()
  final <- bind_rows(final, data.frame(subtype = st, n = n))
}
final
```

## 3mer Spectra - All Subjects

```{r}
get_subtype_3mer <- function(subtype){
  data_dir <- "/net/snowwhite/home/beckandy/research/BRIDGES_redo/output/singletons"
  awk_cmd <- paste0("awk '{count[substr($1,10,3)]++}END{for(key in count)print(key, count[key])}' ",
                    data_dir, "/", subtype, ".txt")
  df <- vroom::vroom(pipe(awk_cmd), col_names = c("Motif", "n"), show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(subtype = str_replace_all(subtype,"_","->")) %>%
    mutate(Motif = paste0(str_sub(Motif, 1, 1), "(", str_remove(subtype, "cpg->"), ")", str_sub(Motif, 3, 3)))
  return(df)
}

get_all_3mer <- function(){
  subtypes <- c("AT_CG", "AT_GC", "AT_TA", "GC_AT", "GC_TA", "GC_CG", "cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG")
  df_list <- vector("list", length(subtypes))
  names(df_list) <- subtypes
  for(st in subtypes){
    df_list[[st]] <- get_subtype_3mer(st)
  }
  return(bind_rows(df_list))
}

bridges_3mer <- get_all_3mer()
```

```{r}
bridges_3mer <- bridges_3mer %>%
  mutate(subtype = str_sub(Motif, 3, 7)) %>%
  ungroup() %>%
  mutate(pct = n / sum(n))
```

```{r}
p <- bridges_3mer %>%
  arrange(subtype, Motif) %>%
  mutate(Motif = factor(Motif, levels=Motif)) %>%
  ggplot(aes(x = Motif, y = pct, fill = subtype, order = subtype)) + 
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab("Subtype") + ylab("Relative Contribution") + ggtitle("BRIDGES Singleton Spectrum")

p

tikz("output/tikz/3mer_spec_BRIDGES.tex", width = 7, height = 3.2)
print(p)
dev.off()
```

