---
title: "null_dist"
author: "Andy Beck"
date: "2022-08-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Here we will be looking at the deviance statistics computed when comparing singleton and control observations 

```{r}
library(tidyverse)
source("code/single_pos_functions.R")

#input_dir <- "/net/snowwhite/home/beckandy/research/1000G_LSCI/output/null_dist/ALL/"
input_dir <- "/net/snowwhite/home/beckandy/research/1000G_LSCI/output/null_dist/no_mask/ALL/"
```

 Let's get the table for a single subtype. For this we will need to aggregate across the 22 autosomes:

```{r}
read_csv(paste0(input_dir, "singleton_9_AT_GC.csv"), show_col_types = FALSE) %>%
  mutate(Nuc = toupper(Nuc)) %>%
  group_by(Nuc) %>%
  summarise(Singletons = sum(Singletons))
```

 
```{r}
load_subtype_singleton_chromosome <- function(subtype, input_dir, chrom){
  f_name <- paste0(input_dir, "singleton_",chrom,"_",subtype,".csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
    mutate(Nuc = toupper(Nuc)) %>%
    group_by(Nuc) %>%
    summarise(Singletons = sum(Singletons)) %>%
    filter(Nuc %in% c("A", "T", "C", "G"))
  return(df)
}

load_subtype_control_chromosome <- function(subtype, input_dir, chrom){
  f_name <- paste0(input_dir, "control_",chrom,"_",subtype,".csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
    mutate(Nuc = toupper(Nuc)) %>%
    group_by(Nuc) %>%
    summarise(Controls = sum(Controls)) %>%
    filter(Nuc %in% c("A", "T", "C", "G"))
  return(df)
}

load_subtype_singletons <- function(subtype, input_dir){
  df <- load_subtype_singleton_chromosome(subtype, input_dir, 1)
  for(i in 2:22){
    df2 <- load_subtype_singleton_chromosome(subtype, input_dir, i)
    df <- full_join(df, df2, by = "Nuc") %>%
      mutate(Singletons = Singletons.x + Singletons.y) %>%
      select(Nuc, Singletons)
  }
  return(df)
}

load_subtype_controls <- function(subtype, input_dir){
  df <- load_subtype_control_chromosome(subtype, input_dir, 1)
  for(i in 2:22){
    df2 <- load_subtype_control_chromosome(subtype, input_dir, i)
    df <- full_join(df, df2, by = "Nuc") %>%
      mutate(Controls = Controls.x + Controls.y) %>%
      select(Nuc, Controls)
  }
  return(df)
}

load_table <- function(subtype, input_dir){
  df_s <- load_subtype_singletons(subtype, input_dir)
  df_c <- load_subtype_controls(subtype, input_dir)
  df <- full_join(df_s, df_c, by = "Nuc") %>%
    arrange(Nuc)
  return(df)
}

re_stat <- function(subtype, input_dir){
  df <- load_table(subtype, input_dir) %>%
    pivot_longer(-Nuc, names_to = "type", values_to = "n") 
  mod_obj <- glm(n ~ Nuc + type, data = df, family = poisson())
  df$re <- 2 * df$n * log(df$n / mod_obj$fitted.values) / (2 * sum(df$n))
  return(sum(df$re))
}

deviance_stat <- function(subtype, input_dir){
  df <- load_table(subtype, input_dir) %>%
    pivot_longer(-Nuc, names_to = "type", values_to = "n")
  mod_obj <- glm(n ~ Nuc + type, data = df, family = poisson())
  final <- list(deviance = deviance(mod_obj), re = deviance(mod_obj) / (2*sum(df$n)))
  return(final)
}
```

Let's look at our subtypes:

```{r}
plot_pos_stat("ALL", "AT_TA", dev_by_position, title_text = "Deviance: ", ylab_text = "Deviance") +
  geom_hline(yintercept = 4.12, colour = "red")

dev_by_position("ALL", "AT_TA")
zhu_by_pos("ALL", "AT_TA")
```

And if we look at the value of the statistic when considering relative position +1000:

```{r}
deviance_stat("AT_TA", input_dir)
```

Let's go ahead and grab the +1000 statistic for each subtype:

```{r}
get_all_long_stat <- function(input_dir){
  subtypes <- c("AT_CG", "AT_GC", "AT_TA",
                "GC_AT", "GC_TA", "GC_CG",
                "cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG")
  dev <- c()
  re <- c()
  for(st in subtypes){
    val <- deviance_stat(st, input_dir)
    dev <- c(dev, val$deviance)
    re <- c(re, val$re)
  }
  return(data.frame(st = subtypes, dev = dev, re = re))
}

get_all_long_stat(input_dir) %>% knitr::kable()
```


