---
title: "Single Position Comparisons"
author: "Andy Beck"
date: "2023-05-31"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)
# library(egg)
library(corrr)
#library(tikzDevice)
#library(sjPlot)
width = 4
height = 4

## Plot options
theme_set(theme_classic())
plot_w <- 800
plot_h <- 500
plot_dpi <- 300
plot_dir <- "output/figs/"

cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")
```

## Introduction

Here in this document we will compare results across the five 1kGP super-populations to see if there are any differences in the patterns of influence of flanking positions on rates of mutation.

### Code for loading results

```{r}
read_subtype_pop_res <- function(pop, subtype, data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/single_pos/"){
  fname <- paste0(data_dir, pop, "/", subtype, ".csv")
  df <- read_csv(fname, show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons + controls)))
  return(df)
}

read_subtype_all_pop <- function(subtype,
                                 pops = c("AFR", "AMR", "EAS", "EUR", "SAS"),
                                 data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/single_pos/"){
  
  results <- lapply(pops, function(x) NULL) %>% setNames(pops)
  for(pop in pops){
    results[[pop]] <- read_subtype_pop_res(pop, subtype, data_dir)
    results[[pop]]$pop <- pop
  }
  df <- bind_rows(results)
  return(df)
}

read_pop_res <- function(pop, data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/single_pos/"){
  df <- read_subtype_pop_res(pop, "AT_CG", data_dir = data_dir)
  df$subtype <- "AT_CG"
  for(st in c("AT_GC", "AT_TA", 
              "GC_AT", "GC_TA", "GC_CG",
              "cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG") ) {
    df2 <- read_subtype_pop_res(pop, st, data_dir = data_dir)
    df2$subtype <- st
    df <- bind_rows(df, df2)
  }
  return(df)
}

read_subtype_pop_resid <- function(pop, subtype, rp, data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/single_pos/resid/"){
  f_name <- paste0(data_dir, pop, "/", subtype, "_rp_", rp, ".csv")
  df <- read_csv(f_name,show_col_types = FALSE)
  return(df)
}

read_subtype_pop_range_resid <- function(pop, subtype, size = 10, data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/single_pos/resid/"){
  if(str_starts(subtype, "cpg")){
    pos.names <- c(-size:-1, 2:size)
  } else{
    pos.names <- c(-size:-1, 1:size)
  }
  results <- lapply(pos.names, function(x) NULL) %>% setNames(pos.names)
  for(i in names(results)){
    results[[i]] <- read_subtype_pop_resid(pop, subtype, i, data_dir)
    results[[i]]$pos <- i
  }
  df <- bind_rows(results)
  return(df)
}

read_subtype_range_resid <- function(subtype,size=10, 
                                     pops = c("AFR", "AMR", "EAS", "EUR", "SAS"),
                                     data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/single_pos/resid/"
){
  results <- lapply(pops, function(x) NULL) %>% setNames(pops)
  for(pop in pops){
    results[[pop]] <- read_subtype_pop_range_resid(pop, subtype, size = size, data_dir)
    results[[pop]]$pop <- pop
  }
  df <- bind_rows(results)
  return(df)
}

subtype_print_names <- function(st){
  if(str_starts(st, "AT")){
    return(paste0("A → ", str_sub(st, 4, 4)))
  } else if(str_starts(st, "GC")){
    return(paste0("C → ", str_sub(st, 5, 5)))
  } else{
    return(paste0("CpG → ", str_sub(st, 9, 9), "pG"))
  }
}
```

## Position-level Results

### AT_CG

```{r}
subtype <- "AT_CG"
df <- read_subtype_all_pop(subtype) 

df %>%
  filter(abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re, colour = pop)) +
  geom_line() +
  geom_point() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  labs(colour = "Population") +
  ggtitle("Single Position Results", subtype)
```

We can also get the correlation of the results between each pair of populations:

```{r}

df %>%
  select(pop, offset, re) %>%
  pivot_wider(names_from = pop, values_from = re) %>%
  select(-offset) %>%
  corrr::correlate()
```

Each position-level result can be further broken down into the contribution from the nucleotide-level residuals:


```{r}
df_resid <- read_subtype_range_resid(subtype) %>%
  type_convert()

df_resid <- df_resid %>%
  inner_join({df %>% select(offset, re, pop)}, by = c("pos" = "offset", "pop"))

df_resid %>% 
  filter(status == "singletons") %>%
  select(nuc, res, re, pos, pop) %>%
  group_by(pop, pos) %>%
  mutate(pct = abs(res) / sum(abs(res))) %>%
  mutate(re_cont = re * pct * sign(res)) %>%
  ggplot(aes(x = pos, y = re_cont, color = nuc)) +
  geom_point() +
  geom_line() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  guides(color = guide_legend(title = "Nucleotide")) +
  theme_classic() +
  theme(text = element_text(family = "Times New Roman"),
        legend.position = c(.85, .2)) +
  facet_wrap(vars(pop))
```

Widen the residual df so we can get pairwise population correlations:

```{r}
df_wide <- df_resid %>%
  filter(status == "singletons") %>%
  select(nuc, pos, pop, res) %>%
  pivot_wider(names_from = pop, values_from = res)
```

And generate the correlations:

```{r}
cor_method <- "pearson"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))
```

And then plot them: 

```{r}
df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

Let's also do Spearman correlations within each position:

```{r}
cor_method <- "spearman"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

Let's take a look at the residuals for AFR and EUR at the +2 position:

```{r}
df_resid %>%
  filter(pop %in% c("AFR", "EUR"),
         pos == 2,
         status == "singletons") %>%
  select(res, pop, nuc) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  ggplot(aes(x = AFR, y=EUR, color = nuc)) +
  geom_point()
```



Trying to do some correlation of the nucleotide residuals:

```{r}
df_resid %>%
  select(pos, nuc, status, res, pop) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  arrange(pos, status, nuc) %>%
  filter(status == "singletons") %>%
  select(AFR, AMR, EAS, EUR, SAS) %>%
  correlate()
```

### AT_GC

```{r}
subtype <- "AT_GC"
df <- read_subtype_all_pop(subtype) 

df %>%
  filter(abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re, colour = pop)) +
  geom_line() +
  geom_point() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  labs(colour = "Population") +
  ggtitle("Single Position Results", subtype)
```

```{r}
df %>%
  select(pop, offset, re) %>%
  pivot_wider(names_from = pop, values_from = re) %>%
  select(-offset) %>%
  corrr::correlate()
```

```{r}
df_resid <- read_subtype_range_resid(subtype) %>%
  type_convert()

df_resid <- df_resid %>%
  inner_join({df %>% select(offset, re, pop)}, by = c("pos" = "offset", "pop"))

df_resid %>% 
  filter(status == "singletons") %>%
  select(nuc, res, re, pos, pop) %>%
  group_by(pop, pos) %>%
  mutate(pct = abs(res) / sum(abs(res))) %>%
  mutate(re_cont = re * pct * sign(res)) %>%
  ggplot(aes(x = pos, y = re_cont, color = nuc)) +
  geom_point() +
  geom_line() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  guides(color = guide_legend(title = "Nucleotide")) +
  theme_classic() +
  theme(text = element_text(family = "Times New Roman"),
        legend.position = c(.85, .2)) +
  facet_wrap(vars(pop))
```

```{r}
df_wide <- df_resid %>%
  filter(status == "singletons") %>%
  select(nuc, pos, pop, res) %>%
  pivot_wider(names_from = pop, values_from = res)
```

```{r}
cor_method <- "pearson"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))
```

```{r}
df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
cor_method <- "spearman"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
df_resid %>%
  select(pos, nuc, status, res, pop) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  arrange(pos, status, nuc) %>%
  filter(status == "singletons") %>%
  select(AFR, AMR, EAS, EUR, SAS) %>%
  correlate()
```

### AT_TA

```{r}
subtype <- "AT_TA"
df <- read_subtype_all_pop(subtype) 

df %>%
  filter(abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re, colour = pop)) +
  geom_line() +
  geom_point() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  labs(colour = "Population") +
  ggtitle("Single Position Results", subtype)
```

```{r}

df %>%
  select(pop, offset, re) %>%
  pivot_wider(names_from = pop, values_from = re) %>%
  select(-offset) %>%
  corrr::correlate()
```

```{r}
df_resid <- read_subtype_range_resid(subtype) %>%
  type_convert()

df_resid <- df_resid %>%
  inner_join({df %>% select(offset, re, pop)}, by = c("pos" = "offset", "pop"))

df_resid %>% 
  filter(status == "singletons") %>%
  select(nuc, res, re, pos, pop) %>%
  group_by(pop, pos) %>%
  mutate(pct = abs(res) / sum(abs(res))) %>%
  mutate(re_cont = re * pct * sign(res)) %>%
  ggplot(aes(x = pos, y = re_cont, color = nuc)) +
  geom_point() +
  geom_line() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  guides(color = guide_legend(title = "Nucleotide")) +
  theme_classic() +
  theme(text = element_text(family = "Times New Roman"),
        legend.position = c(.85, .2)) +
  facet_wrap(vars(pop))
```

```{r}
df_wide <- df_resid %>%
  filter(status == "singletons") %>%
  select(nuc, pos, pop, res) %>%
  pivot_wider(names_from = pop, values_from = res)
```

```{r}
cor_method <- "pearson"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))
```

```{r}
df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
cor_method <- "spearman"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
df_resid %>%
  select(pos, nuc, status, res, pop) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  arrange(pos, status, nuc) %>%
  filter(status == "singletons") %>%
  select(AFR, AMR, EAS, EUR, SAS) %>%
  correlate()
```

### GC_AT

```{r}
subtype <- "GC_AT"
df <- read_subtype_all_pop(subtype) 

df %>%
  filter(abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re, colour = pop)) +
  geom_line() +
  geom_point() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  labs(colour = "Population") +
  ggtitle("Single Position Results", subtype)
```

```{r}
df %>%
  select(pop, offset, re) %>%
  pivot_wider(names_from = pop, values_from = re) %>%
  select(-offset) %>%
  corrr::correlate()
```

```{r}
df_resid <- read_subtype_range_resid(subtype) %>%
  type_convert()

df_resid <- df_resid %>%
  inner_join({df %>% select(offset, re, pop)}, by = c("pos" = "offset", "pop"))

df_resid %>% 
  filter(status == "singletons") %>%
  select(nuc, res, re, pos, pop) %>%
  group_by(pop, pos) %>%
  mutate(pct = abs(res) / sum(abs(res))) %>%
  mutate(re_cont = re * pct * sign(res)) %>%
  ggplot(aes(x = pos, y = re_cont, color = nuc)) +
  geom_point() +
  geom_line() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  guides(color = guide_legend(title = "Nucleotide")) +
  theme_classic() +
  theme(text = element_text(family = "Times New Roman"),
        legend.position = c(.85, .2)) +
  facet_wrap(vars(pop))
```

```{r}
df_wide <- df_resid %>%
  filter(status == "singletons") %>%
  select(nuc, pos, pop, res) %>%
  pivot_wider(names_from = pop, values_from = res)

cor_method <- "pearson"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
cor_method <- "spearman"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
df_resid %>%
  select(pos, nuc, status, res, pop) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  arrange(pos, status, nuc) %>%
  filter(status == "singletons") %>%
  select(AFR, AMR, EAS, EUR, SAS) %>%
  correlate()
```

### GC_TA

```{r}
subtype <- "GC_TA"
df <- read_subtype_all_pop(subtype) 

df %>%
  filter(abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re, colour = pop)) +
  geom_line() +
  geom_point() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  labs(colour = "Population") +
  ggtitle("Single Position Results", subtype)
```

```{r}
df %>%
  select(pop, offset, re) %>%
  pivot_wider(names_from = pop, values_from = re) %>%
  select(-offset) %>%
  corrr::correlate()
```

```{r}
df_resid <- read_subtype_range_resid(subtype) %>%
  type_convert()

df_resid <- df_resid %>%
  inner_join({df %>% select(offset, re, pop)}, by = c("pos" = "offset", "pop"))

df_resid %>% 
  filter(status == "singletons") %>%
  select(nuc, res, re, pos, pop) %>%
  group_by(pop, pos) %>%
  mutate(pct = abs(res) / sum(abs(res))) %>%
  mutate(re_cont = re * pct * sign(res)) %>%
  ggplot(aes(x = pos, y = re_cont, color = nuc)) +
  geom_point() +
  geom_line() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  guides(color = guide_legend(title = "Nucleotide")) +
  theme_classic() +
  theme(text = element_text(family = "Times New Roman"),
        legend.position = c(.85, .2)) +
  facet_wrap(vars(pop))
```

```{r}
df_wide <- df_resid %>%
  filter(status == "singletons") %>%
  select(nuc, pos, pop, res) %>%
  pivot_wider(names_from = pop, values_from = res)

cor_method <- "pearson"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
cor_method <- "spearman"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
df_resid %>%
  select(pos, nuc, status, res, pop) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  arrange(pos, status, nuc) %>%
  filter(status == "singletons") %>%
  select(AFR, AMR, EAS, EUR, SAS) %>%
  correlate()
```

### GC_CG

```{r}
subtype <- "GC_CG"
df <- read_subtype_all_pop(subtype) 

df %>%
  filter(abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re, colour = pop)) +
  geom_line() +
  geom_point() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  labs(colour = "Population") +
  ggtitle("Single Position Results", subtype)
```

```{r}
df %>%
  select(pop, offset, re) %>%
  pivot_wider(names_from = pop, values_from = re) %>%
  select(-offset) %>%
  corrr::correlate()
```

```{r}
df_resid <- read_subtype_range_resid(subtype) %>%
  type_convert()

df_resid <- df_resid %>%
  inner_join({df %>% select(offset, re, pop)}, by = c("pos" = "offset", "pop"))

df_resid %>% 
  filter(status == "singletons") %>%
  select(nuc, res, re, pos, pop) %>%
  group_by(pop, pos) %>%
  mutate(pct = abs(res) / sum(abs(res))) %>%
  mutate(re_cont = re * pct * sign(res)) %>%
  ggplot(aes(x = pos, y = re_cont, color = nuc)) +
  geom_point() +
  geom_line() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  guides(color = guide_legend(title = "Nucleotide")) +
  theme_classic() +
  theme(text = element_text(family = "Times New Roman"),
        legend.position = c(.85, .2)) +
  facet_wrap(vars(pop))
```

```{r}
df_wide <- df_resid %>%
  filter(status == "singletons") %>%
  select(nuc, pos, pop, res) %>%
  pivot_wider(names_from = pop, values_from = res)

cor_method <- "pearson"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
cor_method <- "spearman"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
df_resid %>%
  select(pos, nuc, status, res, pop) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  arrange(pos, status, nuc) %>%
  filter(status == "singletons") %>%
  select(AFR, AMR, EAS, EUR, SAS) %>%
  correlate()
```

### cpg_GC_AT

```{r}
subtype <- "cpg_GC_AT"
df <- read_subtype_all_pop(subtype) 

df %>%
  filter(abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re, colour = pop)) +
  geom_line() +
  geom_point() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  labs(colour = "Population") +
  ggtitle("Single Position Results", subtype)
```

```{r}
df %>%
  select(pop, offset, re) %>%
  pivot_wider(names_from = pop, values_from = re) %>%
  select(-offset) %>%
  corrr::correlate()
```

```{r}
df_resid <- read_subtype_range_resid(subtype) %>%
  type_convert()

df_resid <- df_resid %>%
  inner_join({df %>% select(offset, re, pop)}, by = c("pos" = "offset", "pop"))

df_resid %>% 
  filter(status == "singletons") %>%
  select(nuc, res, re, pos, pop) %>%
  group_by(pop, pos) %>%
  mutate(pct = abs(res) / sum(abs(res))) %>%
  mutate(re_cont = re * pct * sign(res)) %>%
  ggplot(aes(x = pos, y = re_cont, color = nuc)) +
  geom_point() +
  geom_line() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  guides(color = guide_legend(title = "Nucleotide")) +
  theme_classic() +
  theme(text = element_text(family = "Times New Roman"),
        legend.position = c(.85, .2)) +
  facet_wrap(vars(pop))
```

```{r}
df_wide <- df_resid %>%
  filter(status == "singletons") %>%
  select(nuc, pos, pop, res) %>%
  pivot_wider(names_from = pop, values_from = res)

cor_method <- "pearson"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
cor_method <- "spearman"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
df_resid %>%
  select(pos, nuc, status, res, pop) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  arrange(pos, status, nuc) %>%
  filter(status == "singletons") %>%
  select(AFR, AMR, EAS, EUR, SAS) %>%
  correlate()
```

### cpg_GC_TA

```{r}
subtype <- "cpg_GC_TA"
df <- read_subtype_all_pop(subtype) 

df %>%
  filter(abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re, colour = pop)) +
  geom_line() +
  geom_point() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  labs(colour = "Population") +
  ggtitle("Single Position Results", subtype)
```

```{r}
df %>%
  select(pop, offset, re) %>%
  pivot_wider(names_from = pop, values_from = re) %>%
  select(-offset) %>%
  corrr::correlate()
```

```{r}
df_resid <- read_subtype_range_resid(subtype) %>%
  type_convert()

df_resid <- df_resid %>%
  inner_join({df %>% select(offset, re, pop)}, by = c("pos" = "offset", "pop"))

df_resid %>% 
  filter(status == "singletons") %>%
  select(nuc, res, re, pos, pop) %>%
  group_by(pop, pos) %>%
  mutate(pct = abs(res) / sum(abs(res))) %>%
  mutate(re_cont = re * pct * sign(res)) %>%
  ggplot(aes(x = pos, y = re_cont, color = nuc)) +
  geom_point() +
  geom_line() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  guides(color = guide_legend(title = "Nucleotide")) +
  theme_classic() +
  theme(text = element_text(family = "Times New Roman"),
        legend.position = c(.85, .2)) +
  facet_wrap(vars(pop))
```

```{r}
df_wide <- df_resid %>%
  filter(status == "singletons") %>%
  select(nuc, pos, pop, res) %>%
  pivot_wider(names_from = pop, values_from = res)

cor_method <- "pearson"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
cor_method <- "spearman"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
df_resid %>%
  select(pos, nuc, status, res, pop) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  arrange(pos, status, nuc) %>%
  filter(status == "singletons") %>%
  select(AFR, AMR, EAS, EUR, SAS) %>%
  correlate()
```

### cpg_GC_CG

```{r}
subtype <- "cpg_GC_CG"
df <- read_subtype_all_pop(subtype) 

df %>%
  filter(abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re, colour = pop)) +
  geom_line() +
  geom_point() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  labs(colour = "Population") +
  ggtitle("Single Position Results", subtype)
```

```{r}
df %>%
  select(pop, offset, re) %>%
  pivot_wider(names_from = pop, values_from = re) %>%
  select(-offset) %>%
  corrr::correlate()
```

```{r}
df_resid <- read_subtype_range_resid(subtype) %>%
  type_convert()

df_resid <- df_resid %>%
  inner_join({df %>% select(offset, re, pop)}, by = c("pos" = "offset", "pop"))

df_resid %>% 
  filter(status == "singletons") %>%
  select(nuc, res, re, pos, pop) %>%
  group_by(pop, pos) %>%
  mutate(pct = abs(res) / sum(abs(res))) %>%
  mutate(re_cont = re * pct * sign(res)) %>%
  ggplot(aes(x = pos, y = re_cont, color = nuc)) +
  geom_point() +
  geom_line() +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  guides(color = guide_legend(title = "Nucleotide")) +
  theme_classic() +
  theme(text = element_text(family = "Times New Roman"),
        legend.position = c(.85, .2)) +
  facet_wrap(vars(pop))
```

```{r}
df_wide <- df_resid %>%
  filter(status == "singletons") %>%
  select(nuc, pos, pop, res) %>%
  pivot_wider(names_from = pop, values_from = res)

cor_method <- "pearson"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
cor_method <- "spearman"
df_cor <- df_wide %>%
    group_by(pos) %>%
    summarise(across(AMR:SAS, ~cor(., AFR, use = "complete.obs", method = cor_method), .names = "AFR_{.col}"),
              across(EAS:SAS, ~cor(., AMR, use = "complete.obs", method = cor_method), .names = "AMR_{.col}"),
              across(EUR:SAS, ~cor(., EAS, use = "complete.obs", method = cor_method), .names = "EAS_{.col}"),
              across(SAS, ~cor(., EUR, use = "complete.obs", method = cor_method), .names = "EUR_{.col}"))

df_cor %>%
    select(pos, starts_with("AFR")) %>%
    pivot_longer(-pos, names_to = "pair", values_to = "cor") %>%
    mutate(pair = str_remove(pair, "AFR_")) %>%
    ggplot(aes(x = pos, y = cor, group_by = pair, colour = pair)) +
    geom_point() + 
    geom_line() +
    ylim(c(0,1)) +
    ggtitle("Nucleotide residual correlations", 
            paste0("Reference Population: AFR; subtype: ", subtype_print_names(subtype) )) +
    xlab("Relative Position") +
    ylab("Nucleotide Contribution Correlation") +
    labs(colour = "Population")
```

```{r}
df_resid %>%
  select(pos, nuc, status, res, pop) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  arrange(pos, status, nuc) %>%
  filter(status == "singletons") %>%
  select(AFR, AMR, EAS, EUR, SAS) %>%
  correlate()
```

## Chi-sq Approach

Above we looked at the consistency of results for the models in which we assess the influence of nucleotides at individual positions. A similar question we can ask is if the distributions of nucleotides flanking singletons from one population are similar to the corresponding distribution in another population. To address this, we employ a $\chi^2$ approach wherein we use the distribution of nucleotides at a flanking position for one population as the expectation for the corresponding distribution for another population.

### Loading Results

```{r}
chi_sq_dir <- "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/ref_pop_sp/"
load_chi_res <- function(src, ref, subtype, res_dir){
  f_name <- paste0(res_dir, src, "_", ref, "_", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE)
  return(df)
}
```

### Results

#### AFR as Source Population

##### EUR

```{r}
src_pop <- "AFR"
trg_pop <- "EUR"

subtype <- "AT_CG"

df <- load_chi_res(src_pop, trg_pop, subtype, chi_sq_dir)
df %>%
  head()

df %>%
  ggplot(aes(x = rp, y = stat)) +
  geom_point() +
  xlab("Relative Position") +
  ylab("Chi-Sq Statistic")
```

