---
title: "NYGC_rank_re_vals"
author: "Andy Beck"
date: "2023-06-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(reactable)

## Plot options
theme_set(theme_classic())
plot_w <- 800
plot_h <- 500
plot_dpi <- 300
plot_dir <- "output/figs/"

cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")
```

## Introduction

The purpose of this document is to collect and organize the relative entropy values for the single and two position models. Primarily we are interested in noting where two-position re values are higher than single position model re values.

### Code for reading in results

```{r}
read_single_subtype_pop_res <- function(pop, subtype, 
                                        data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/single_pos/"){
  fname <- paste0(data_dir, pop, "/", subtype, ".csv")
  df <- read_csv(fname, show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons + controls)))
  return(df)
}

read_two_subtype_pop_res <- function(pop, subtype, 
                                        data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/two_pos/"){
  fname <- paste0(data_dir, pop, "/", subtype, ".csv")
  df <- read_csv(fname, show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons + controls)))
  return(df)
}

read_three_subtype_pop_res <- function(pop, subtype, 
                                        data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/three_pos/"){
  fname <- paste0(data_dir, pop, "/", subtype, ".csv")
  df <- read_csv(fname, show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons + controls)))
  return(df)
}

read_both_subtype_pop_res <- function(pop, subtype, 
                                      single_dir="/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/single_pos/",
                                      two_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/two_pos/"){
  df_s <- read_single_subtype_pop_res(pop, subtype, single_dir) %>%
    select(offset, dev, re) %>%
    filter(abs(offset) < 21) %>%
    mutate(offset = as.character(offset))
  
  df_t <- read_two_subtype_pop_res(pop, subtype, two_dir) %>%
    select(rp1, rp2, dev, re) %>%
    rowwise() %>%
    mutate(offset = paste0(rp1, ", ", rp2)) %>%
    select(offset, dev, re)
  df <- bind_rows(df_s, df_t) %>%
    arrange(desc(re))
  return(df)
}

read_one_two_three <- function(pop, subtype, 
                                      single_dir="/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/single_pos/",
                                      two_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/two_pos/",
                               three_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/three_pos/"){
  df_s <- read_single_subtype_pop_res(pop, subtype, single_dir) %>%
    select(offset, dev, re) %>%
    filter(abs(offset) < 21) %>%
    mutate(offset = as.character(offset))
  
  df_t <- read_two_subtype_pop_res(pop, subtype, two_dir) %>%
    select(rp1, rp2, dev, re) %>%
    rowwise() %>%
    mutate(offset = paste0(rp1, ", ", rp2)) %>%
    select(offset, dev, re)
  
  df_trip <- read_three_subtype_pop_res(pop, subtype, three_dir) %>%
    select(rp1, rp2, rp3, dev, re) %>%
    rowwise() %>%
    mutate(offset = paste0(rp1, ", ", rp2, ", ", rp3)) %>%
    select(offset, dev, re)
  
  
  df_s$type <- "Single"
  df_t$type <- "Two-Way"
  df_trip$type <- "Three-Way"
  
  df <- bind_rows(df_s, df_t) %>%
    bind_rows(df_trip) %>%
    arrange(desc(re))
  return(df)
}
```

## Subtype Comparisons

### A to C

```{r}
subtype <- "AT_CG"
df <- read_one_two_three("ALL", subtype)

max_re <- max(df$re)
df$pct_max <- df$re / max_re
df$rank <- rank(-df$re)
df <- df %>%
  group_by(type) %>%
  mutate(class_rank = rank(-re))

df %>% reactable()
```

An attempt at a plot, if you might indulge me:

```{r}
df %>%
  ggplot(aes(x = class_rank, 
             y = pct_max, 
             color = type, 
             shape = type)) +
  geom_point(size = 2.5) +
  scale_colour_manual(values=cbPalette, name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_shape_discrete(name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_x_log10() +
  geom_abline(slope = 0, intercept = 0.05) +
  theme(legend.position=c(.8, .8)) +
  xlab("% of Max RE") +
  ylab("Within-class Rank")
```



### A to G

```{r}
subtype <- "AT_GC"
df <- read_one_two_three("ALL", subtype)

max_re <- max(df$re)
df$pct_max <- df$re / max_re
df$rank <- rank(-df$re)
df <- df %>%
  group_by(type) %>%
  mutate(class_rank = rank(-re))

df %>% reactable()
```

```{r}
df %>%
  ggplot(aes(x = class_rank, 
             y = pct_max, 
             color = type, 
             shape = type)) +
  geom_point(size = 2.5) +
  scale_colour_manual(values=cbPalette, name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_shape_discrete(name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_x_log10() +
  geom_abline(slope = 0, intercept = 0.05) +
  theme(legend.position=c(.8, .8)) +
  xlab("% of Max RE") +
  ylab("Within-class Rank")
```

### A to T

```{r}
subtype <- "AT_TA"
df <- read_one_two_three("ALL", subtype)

max_re <- max(df$re)
df$pct_max <- df$re / max_re
df$rank <- rank(-df$re)
df <- df %>%
  group_by(type) %>%
  mutate(class_rank = rank(-re))

df %>% reactable()
```

```{r}
df %>%
  ggplot(aes(x = class_rank, 
             y = pct_max, 
             color = type, 
             shape = type)) +
  geom_point(size = 2.5) +
  scale_colour_manual(values=cbPalette, name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_shape_discrete(name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_x_log10() +
  geom_abline(slope = 0, intercept = 0.05) +
  theme(legend.position=c(.8, .8)) +
  xlab("% of Max RE") +
  ylab("Within-class Rank")
```

### C to A

```{r}
subtype <- "GC_TA"
df <- read_one_two_three("ALL", subtype)

max_re <- max(df$re)
df$pct_max <- df$re / max_re
df$rank <- rank(-df$re)
df <- df %>%
  group_by(type) %>%
  mutate(class_rank = rank(-re))

df %>% reactable()
```

```{r}
df %>%
  ggplot(aes(x = class_rank, 
             y = pct_max, 
             color = type, 
             shape = type)) +
  geom_point(size = 2.5) +
  scale_colour_manual(values=cbPalette, name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_shape_discrete(name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_x_log10() +
  geom_abline(slope = 0, intercept = 0.05) +
  theme(legend.position=c(.8, .8)) +
  xlab("% of Max RE") +
  ylab("Within-class Rank")
```

### CpG to ApG

```{r}
subtype <- "cpg_GC_TA"
df <- read_one_two_three("ALL", subtype)

max_re <- max(df$re)
df$pct_max <- df$re / max_re
df$rank <- rank(-df$re)
df <- df %>%
  group_by(type) %>%
  mutate(class_rank = rank(-re))

df %>% reactable()
```

```{r}
df %>%
  ggplot(aes(x = class_rank, 
             y = pct_max, 
             color = type, 
             shape = type)) +
  geom_point(size = 2.5) +
  scale_colour_manual(values=cbPalette, name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_shape_discrete(name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_x_log10() +
  geom_abline(slope = 0, intercept = 0.05) +
  theme(legend.position=c(.8, .8)) +
  xlab("% of Max RE") +
  ylab("Within-class Rank")
```

### C to T

```{r}
subtype <- "GC_AT"
df <- read_one_two_three("ALL", subtype)

max_re <- max(df$re)
df$pct_max <- df$re / max_re
df$rank <- rank(-df$re)
df <- df %>%
  group_by(type) %>%
  mutate(class_rank = rank(-re))

df %>% reactable()
```

```{r}
df %>%
  ggplot(aes(x = class_rank, 
             y = pct_max, 
             color = type, 
             shape = type)) +
  geom_point(size = 2.5) +
  scale_colour_manual(values=cbPalette, name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_shape_discrete(name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_x_log10() +
  geom_abline(slope = 0, intercept = 0.05) +
  theme(legend.position=c(.8, .8)) +
  xlab("% of Max RE") +
  ylab("Within-class Rank")
```

### CpG to TpG

```{r}
subtype <- "cpg_GC_AT"
df <- read_one_two_three("ALL", subtype)

max_re <- max(df$re)
df$pct_max <- df$re / max_re
df$rank <- rank(-df$re)
df <- df %>%
  group_by(type) %>%
  mutate(class_rank = rank(-re))

df %>% reactable()
```

```{r}
df %>%
  ggplot(aes(x = class_rank, 
             y = pct_max, 
             color = type, 
             shape = type)) +
  geom_point(size = 2.5) +
  scale_colour_manual(values=cbPalette, name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_shape_discrete(name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_x_log10() +
  geom_abline(slope = 0, intercept = 0.05) +
  theme(legend.position=c(.8, .8)) +
  xlab("% of Max RE") +
  ylab("Within-class Rank")
```

### C to G

```{r}
subtype <- "GC_CG"
df <- read_one_two_three("ALL", subtype)

max_re <- max(df$re)
df$pct_max <- df$re / max_re
df$rank <- rank(-df$re)
df <- df %>%
  group_by(type) %>%
  mutate(class_rank = rank(-re))

df %>% reactable()
```

```{r}
df %>%
  ggplot(aes(x = class_rank, 
             y = pct_max, 
             color = type, 
             shape = type)) +
  geom_point(size = 2.5) +
  scale_colour_manual(values=cbPalette, name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_shape_discrete(name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_x_log10() +
  geom_abline(slope = 0, intercept = 0.05) +
  theme(legend.position=c(.8, .8)) +
  xlab("% of Max RE") +
  ylab("Within-class Rank")
```

### CpG to GpG

```{r}
subtype <- "cpg_GC_CG"
df <- read_one_two_three("ALL", subtype)

max_re <- max(df$re)
df$pct_max <- df$re / max_re
df$rank <- rank(-df$re)
df <- df %>%
  group_by(type) %>%
  mutate(class_rank = rank(-re))

df %>% reactable()
```

```{r}
df %>%
  ggplot(aes(x = class_rank, 
             y = pct_max, 
             color = type, 
             shape = type)) +
  geom_point(size = 2.5) +
  scale_colour_manual(values=cbPalette, name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_shape_discrete(name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
  scale_x_log10() +
  geom_abline(slope = 0, intercept = 0.05) +
  theme(legend.position=c(.8, .8)) +
  xlab("% of Max RE") +
  ylab("Within-class Rank")
```
