---
title: "NYGC_two_pos_comp"
author: "Andy Beck"
date: "2023-06-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(corrr)
library(reshape2)
library(sjPlot)

subtype_print_names <- function(st){
  if(str_starts(st, "AT")){
    return(paste0("A → ", str_sub(st, 4, 4)))
  } else if(str_starts(st, "GC")){
    return(paste0("C → ", str_sub(st, 5, 5)))
  } else{
    return(paste0("CpG → ", str_sub(st, 9, 9), "pG"))
  }
}

#source("code/two_pos_new.R")
```

## Position-level results

Here in this section of the document we will assess the correlation of position-level results across pairs of populations for all of the mutation subtypes.

### Get position-level statistic for each superpopulation

```{r}
get_pos <- function(subtype, 
                    data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/two_pos/"){
  f_name <- paste0(data_dir, "AFR/", subtype, ".csv")
  final <- read_csv(f_name, show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(re = dev / (2*(singletons + controls))) %>%
    select(rp1, rp2, re) %>%
    rename(AFR = re)
  for(pop in c("AMR", "EAS", "EUR", "SAS")){
    f_name <- paste0(data_dir, pop, "/", subtype, ".csv")
    df <- read_csv(f_name, show_col_types = FALSE) %>%
      rowwise() %>%
      mutate(re = dev / (2*(singletons + controls))) %>%
      select(rp1, rp2, re)
    colnames(df)[3] <- pop
    final <- full_join(final, df, by = c("rp1", "rp2"))
  }
  return(final)
}
```

### AT_CG

```{r}
st <- "AT_CG"

simple_cor <- function(st){
  df <- get_pos(st)
  
  cor_mat <- cor({df %>% select(-rp1, -rp2)})
  
  if(!str_starts(st, "cpg")){
    subtype2 <- str_replace(st, "_", " → ")
  } else {
    subtype <- str_sub(st, 5)
    subtype2 <- paste0("(cpg) ", str_replace(subtype, "_", " → "))
  }
  
  p <- cor({df %>% select(-rp1, -rp2)}) %>% 
    melt %>%
    ggplot(aes(x=Var1, y=Var2, fill=value)) +
    geom_tile() +
    xlab("Population") + ylab("Population") + 
    ggtitle("Correlation of Position-Level Results", paste0("Subtype: ", subtype2)) +
    scale_fill_distiller(palette = "Reds", direction = 1, limits = c(0,1))
  return(list(p = p, cor_mat = cor_mat))
}
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### AT_GC

```{r}
st <- "AT_GC"
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### AT_TA

```{r}
st <- "AT_TA"
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### GC_AT

```{r}
st <- "GC_AT"
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### cpg_GC_AT

```{r}
st <- "cpg_GC_AT"
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### GC_TA

```{r}
st <- "GC_TA"
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### cpg_GC_TA

```{r}
st <- "cpg_GC_TA"
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### GC_CG

```{r}
st <- "GC_CG"
obj <-  simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```

### cpg_GC_CG

```{r}
st <- "cpg_GC_CG"
obj <- simple_cor(st)

obj$cor_mat %>% knitr::kable()
obj$p
```


## Do populations share location of highest RE?

```{r}
get_table_re <- function(pop,
                         data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/two_pos/"){
  
  final <- data.frame(st = character(),
                      p1 = numeric(),
                      p2 = numeric(),
                      statistic = numeric())
  
  for(st in c("AT_CG", "AT_GC", "AT_TA", 
              "GC_AT", "GC_TA", "GC_CG",
              "cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG")){
    f_name <- paste0(data_dir, pop, "/", st, ".csv")
    df <- read_csv(f_name, show_col_types = FALSE) %>%
      rowwise() %>%
      mutate(re = dev / (2* (singletons + controls))) %>%
      arrange(desc(re))
    p1 <- df$rp1[1]
    p2 <- df$rp2[1]
    statistic <- df$re[1]
    final <- bind_rows(final,
                       data.frame(st = st,
                                  p1 = p1,
                                  p2 = p2,
                                  statistic = statistic))
  }
  final$pop <- pop
  return(final)
}
```
### ALL

```{r}
pop <- "ALL"
get_table_re(pop) %>%
  knitr::kable()
```

### AFR

```{r}
pop <- "AFR"
get_table_re(pop) %>%
  knitr::kable()
```

### AMR

```{r}
pop <- "AMR"
get_table_re(pop) %>%
  knitr::kable()
```

### EAS

```{r}
pop <- "EAS"
get_table_re(pop) %>%
  knitr::kable()
```

### EUR

```{r}
pop <- "EUR"
get_table_re(pop) %>%
  knitr::kable()
```

### SAS

```{r}
pop <- "SAS"
get_table_re(pop) %>%
  knitr::kable()
```

## GC_AT Differences

```{r}
data_dir <- "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/two_pos/"
pop <- "ALL"
read_csv(paste0(data_dir, pop, "/GC_AT.csv"), show_col_types = FALSE) %>%
  rowwise() %>%
  mutate(re = dev / (2*(singletons + controls))) %>%
  arrange(desc(re)) %>%
  head(4) %>%
  knitr::kable()

pop <- "AFR"
read_csv(paste0(data_dir, pop, "/GC_AT.csv"), show_col_types = FALSE) %>%
  rowwise() %>%
  mutate(re = dev / (2*(singletons + controls))) %>%
  arrange(desc(re)) %>%
  head(4) %>%
  knitr::kable()

pop <- "AMR"
read_csv(paste0(data_dir, pop, "/GC_AT.csv"), show_col_types = FALSE) %>%
  rowwise() %>%
  mutate(re = dev / (2*(singletons + controls))) %>%
  arrange(desc(re)) %>%
  head(4) %>%
  knitr::kable()

pop <- "EAS"
read_csv(paste0(data_dir, pop, "/GC_AT.csv"), show_col_types = FALSE) %>%
  rowwise() %>%
  mutate(re = dev / (2*(singletons + controls))) %>%
  arrange(desc(re)) %>%
  head(4) %>%
  knitr::kable()

pop <- "EUR"
read_csv(paste0(data_dir, pop, "/GC_AT.csv"), show_col_types = FALSE) %>%
  rowwise() %>%
  mutate(re = dev / (2*(singletons + controls))) %>%
  arrange(desc(re)) %>%
  head(4) %>%
  knitr::kable()

pop <- "SAS"
read_csv(paste0(data_dir, pop, "/GC_AT.csv"), show_col_types = FALSE) %>%
  rowwise() %>%
  mutate(re = dev / (2*(singletons + controls))) %>%
  arrange(desc(re)) %>%
  head(4) %>%
  knitr::kable()
```

In this example, while not all populations share the positions coordinates of the highest RE value, we see that (-2, -1) and (-1, 1) are in the top two for all populations.

## AT_CG example (EAS)

For this subtype the only super population to not have the highest observed RE value at (1,2) was EAS (-2, -1)

```{r}
pop <- "ALL"
read_csv(paste0(data_dir, pop, "/AT_CG.csv"), show_col_types = FALSE) %>%
  rowwise() %>%
  mutate(re = dev / (2*(singletons + controls))) %>%
  arrange(desc(re)) %>%
  head(4) %>%
  knitr::kable()

pop <- "EAS"
read_csv(paste0(data_dir, pop, "/AT_CG.csv"), show_col_types = FALSE) %>%
  rowwise() %>%
  mutate(re = dev / (2*(singletons + controls))) %>%
  arrange(desc(re)) %>%
  head(4) %>%
  knitr::kable()

pop <- "AFR"
read_csv(paste0(data_dir, pop, "/AT_CG.csv"), show_col_types = FALSE) %>%
  rowwise() %>%
  mutate(re = dev / (2*(singletons + controls))) %>%
  arrange(desc(re)) %>%
  head(4) %>%
  knitr::kable()

pop <- "AMR"
read_csv(paste0(data_dir, pop, "/AT_CG.csv"), show_col_types = FALSE) %>%
  rowwise() %>%
  mutate(re = dev / (2*(singletons + controls))) %>%
  arrange(desc(re)) %>%
  head(4) %>%
  knitr::kable()

pop <- "EUR"
read_csv(paste0(data_dir, pop, "/AT_CG.csv"), show_col_types = FALSE) %>%
  rowwise() %>%
  mutate(re = dev / (2*(singletons + controls))) %>%
  arrange(desc(re)) %>%
  head(4) %>%
  knitr::kable()

pop <- "SAS"
read_csv(paste0(data_dir, pop, "/AT_CG.csv"), show_col_types = FALSE) %>%
  rowwise() %>%
  mutate(re = dev / (2*(singletons + controls))) %>%
  arrange(desc(re)) %>%
  head(4) %>%
  knitr::kable()
```

Again, we see that across all populations the top four statistics are at the same pairs of positions, with only the order slightly adjusted.


## Residual-level Comparisons

```{r}
read_subtype_pop_resid <- function(pop, subtype, rp1, rp2, data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/two_pos/resid/"){
  f_name <- paste0(data_dir, pop, "/", subtype, "_p", rp1, "_q", rp2, ".csv")
  df <- read_csv(f_name,show_col_types = FALSE)
  return(df)
}

read_subtype_pop_range_resid <- function(pop, subtype, size = 10, data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/two_pos/resid/"){
  if(str_starts(subtype, "cpg")){
    pos.list <- c(-size:-1, 2:size)
  } else{
    pos.list <- c(-size:-1, 1:size)
  }
  pos.names <- c()
  for(i in 1:(length(pos.list) - 1)){
    for(j in (i+1):length(pos.list)){
      pos.names <- c(pos.names, paste0(pos.list[[i]], "_", pos.list[[j]]))
    }
  }
  results <- lapply(pos.names, function(x) NULL) %>% setNames(pos.names)
  for(i in names(results)){
    rp1 <- str_split(i, "_")[[1]][1] %>% as.numeric()
    rp2 <- str_split(i, "_")[[1]][2] %>% as.numeric()
    results[[i]] <- read_subtype_pop_resid(pop, subtype, rp1, rp2, data_dir)
    results[[i]]$rp1 <- rp1
    results[[i]]$rp2 <- rp2
  }
  df <- bind_rows(results)
  return(df)
}

read_subtype_range_resid <- function(subtype, size = 10, data_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/two_pos/resid/"){
  df <- read_subtype_pop_range_resid("ALL", subtype, size = size, data_dir = data_dir)
  df$pop <- "ALL"
  
  for(pop in c("AFR", "AMR", "EAS", "EUR", "SAS")){
    df2 <- read_subtype_pop_range_resid(pop, subtype, size = size, data_dir = data_dir)
    df2$pop <- pop
    df <- bind_rows(df, df2)
  }
  
  return(df)
}
```

### A > C

```{r}
df <- read_subtype_range_resid("AT_CG")

df_cor <- df %>%
  select(pop, status, rp1, rp2, p1, p2, res) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  select_if(is.numeric) %>%
  group_by(rp1, rp2) %>%
  nest() %>%
   mutate(
    correlations = map(data, \(x) correlate(x, quiet = T))
  ) %>%
  unnest(correlations) %>%
  select(-data)
```

```{r}
pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")

for(i in 1:4){
  for(j in (i+1):5){
    p <- df_cor %>%
      filter(term == pops[i]) %>%
      select(rp1, rp2, !!sym(pops[j])) %>%
      ggplot(aes(x = rp2, y = rp1, fill = !!sym(pops[j]))) +
      geom_tile() +
      scale_fill_distiller(palette = "RdBu", limits = c(-1,1)) +
      theme_bw() +
      xlab("Relative Position") +
      ylab("Relative Position") +
      labs(fill = "Residual \nCorrelation") +
      theme(legend.position = c(.1, .8)) +
      annotate("text", x = 0, y = 9, label= paste0(pops[i], " - ", pops[j]))
    print(p)
  }
}
```

### A > G

```{r, message=FALSE, warning=FALSE}
subtype <- "AT_GC"
df <- read_subtype_range_resid(subtype)

df_cor <- df %>%
  select(pop, status, rp1, rp2, p1, p2, res) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  select_if(is.numeric) %>%
  group_by(rp1, rp2) %>%
  nest() %>%
   mutate(
    correlations = map(data, \(x) correlate(x, quiet = T))
  ) %>%
  unnest(correlations) %>%
  select(-data)

pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")

for(i in 1:4){
  for(j in (i+1):5){
    p <- df_cor %>%
      filter(term == pops[i]) %>%
      select(rp1, rp2, !!sym(pops[j])) %>%
      ggplot(aes(x = rp2, y = rp1, fill = !!sym(pops[j]))) +
      geom_tile() +
      scale_fill_distiller(palette = "RdBu", limits = c(-1,1)) +
      theme_bw() +
      xlab("Relative Position") +
      ylab("Relative Position") +
      labs(fill = "Residual \nCorrelation") +
      theme(legend.position = c(.1, .8)) +
      annotate("text", x = 0, y = 9, label= paste0(pops[i], " - ", pops[j]))
    print(p)
  }
}
```

### A > T

```{r, message=FALSE, warning=FALSE}
subtype <- "AT_TA"
df <- read_subtype_range_resid(subtype)

df_cor <- df %>%
  select(pop, status, rp1, rp2, p1, p2, res) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  select_if(is.numeric) %>%
  group_by(rp1, rp2) %>%
  nest() %>%
   mutate(
    correlations = map(data, \(x) correlate(x, quiet = T))
  ) %>%
  unnest(correlations) %>%
  select(-data)

pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")

for(i in 1:4){
  for(j in (i+1):5){
    p <- df_cor %>%
      filter(term == pops[i]) %>%
      select(rp1, rp2, !!sym(pops[j])) %>%
      ggplot(aes(x = rp2, y = rp1, fill = !!sym(pops[j]))) +
      geom_tile() +
      scale_fill_distiller(palette = "RdBu", limits = c(-1,1)) +
      theme_bw() +
      xlab("Relative Position") +
      ylab("Relative Position") +
      labs(fill = "Residual \nCorrelation") +
      theme(legend.position = c(.1, .8)) +
      annotate("text", x = 0, y = 9, label= paste0(pops[i], " - ", pops[j]))
    print(p)
  }
}
```

### C > A

```{r, message=FALSE, warning=FALSE}
subtype <- "GC_TA"
df <- read_subtype_range_resid(subtype)

df_cor <- df %>%
  select(pop, status, rp1, rp2, p1, p2, res) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  select_if(is.numeric) %>%
  group_by(rp1, rp2) %>%
  nest() %>%
   mutate(
    correlations = map(data, \(x) correlate(x, quiet = T))
  ) %>%
  unnest(correlations) %>%
  select(-data)

pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")

for(i in 1:4){
  for(j in (i+1):5){
    p <- df_cor %>%
      filter(term == pops[i]) %>%
      select(rp1, rp2, !!sym(pops[j])) %>%
      ggplot(aes(x = rp2, y = rp1, fill = !!sym(pops[j]))) +
      geom_tile() +
      scale_fill_distiller(palette = "RdBu", limits = c(-1,1)) +
      theme_bw() +
      xlab("Relative Position") +
      ylab("Relative Position") +
      labs(fill = "Residual \nCorrelation") +
      theme(legend.position = c(.1, .8)) +
      annotate("text", x = 0, y = 9, label= paste0(pops[i], " - ", pops[j]))
    print(p)
  }
}
```

### C > T

```{r, message=FALSE, warning=FALSE}
subtype <- "GC_AT"
df <- read_subtype_range_resid(subtype)

df_cor <- df %>%
  select(pop, status, rp1, rp2, p1, p2, res) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  select_if(is.numeric) %>%
  group_by(rp1, rp2) %>%
  nest() %>%
   mutate(
    correlations = map(data, \(x) correlate(x, quiet = T))
  ) %>%
  unnest(correlations) %>%
  select(-data)

pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")

for(i in 1:4){
  for(j in (i+1):5){
    p <- df_cor %>%
      filter(term == pops[i]) %>%
      select(rp1, rp2, !!sym(pops[j])) %>%
      ggplot(aes(x = rp2, y = rp1, fill = !!sym(pops[j]))) +
      geom_tile() +
      scale_fill_distiller(palette = "RdBu", limits = c(-1,1)) +
      theme_bw() +
      xlab("Relative Position") +
      ylab("Relative Position") +
      labs(fill = "Residual \nCorrelation") +
      theme(legend.position = c(.1, .8)) +
      annotate("text", x = 0, y = 9, label= paste0(pops[i], " - ", pops[j]))
    print(p)
  }
}
```

### C > G

```{r, message=FALSE, warning=FALSE}
subtype <- "GC_CG"
df <- read_subtype_range_resid(subtype)

df_cor <- df %>%
  select(pop, status, rp1, rp2, p1, p2, res) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  select_if(is.numeric) %>%
  group_by(rp1, rp2) %>%
  nest() %>%
   mutate(
    correlations = map(data, \(x) correlate(x, quiet = T))
  ) %>%
  unnest(correlations) %>%
  select(-data)

pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")

for(i in 1:4){
  for(j in (i+1):5){
    p <- df_cor %>%
      filter(term == pops[i]) %>%
      select(rp1, rp2, !!sym(pops[j])) %>%
      ggplot(aes(x = rp2, y = rp1, fill = !!sym(pops[j]))) +
      geom_tile() +
      scale_fill_distiller(palette = "RdBu", limits = c(-1,1)) +
      theme_bw() +
      xlab("Relative Position") +
      ylab("Relative Position") +
      labs(fill = "Residual \nCorrelation") +
      theme(legend.position = c(.1, .8)) +
      annotate("text", x = 0, y = 9, label= paste0(pops[i], " - ", pops[j]))
    print(p)
  }
}
```

### CpG > TpG

```{r, message=FALSE, warning=FALSE}
subtype <- "cpg_GC_AT"
df <- read_subtype_range_resid(subtype)

df_cor <- df %>%
  select(pop, status, rp1, rp2, p1, p2, res) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  select_if(is.numeric) %>%
  group_by(rp1, rp2) %>%
  nest() %>%
   mutate(
    correlations = map(data, \(x) correlate(x, quiet = T))
  ) %>%
  unnest(correlations) %>%
  select(-data)

pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")

for(i in 1:4){
  for(j in (i+1):5){
    p <- df_cor %>%
      filter(term == pops[i]) %>%
      select(rp1, rp2, !!sym(pops[j])) %>%
      ggplot(aes(x = rp2, y = rp1, fill = !!sym(pops[j]))) +
      geom_tile() +
      scale_fill_distiller(palette = "RdBu", limits = c(-1,1)) +
      theme_bw() +
      xlab("Relative Position") +
      ylab("Relative Position") +
      labs(fill = "Residual \nCorrelation") +
      theme(legend.position = c(.1, .8)) +
      annotate("text", x = 0, y = 9, label= paste0(pops[i], " - ", pops[j]))
    print(p)
  }
}
```

### CpG > ApG

```{r, message=FALSE, warning=FALSE}
subtype <- "cpg_GC_TA"
df <- read_subtype_range_resid(subtype)

df_cor <- df %>%
  select(pop, status, rp1, rp2, p1, p2, res) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  select_if(is.numeric) %>%
  group_by(rp1, rp2) %>%
  nest() %>%
   mutate(
    correlations = map(data, \(x) correlate(x, quiet = T))
  ) %>%
  unnest(correlations) %>%
  select(-data)

pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")

for(i in 1:4){
  for(j in (i+1):5){
    p <- df_cor %>%
      filter(term == pops[i]) %>%
      select(rp1, rp2, !!sym(pops[j])) %>%
      ggplot(aes(x = rp2, y = rp1, fill = !!sym(pops[j]))) +
      geom_tile() +
      scale_fill_distiller(palette = "RdBu", limits = c(-1,1)) +
      theme_bw() +
      xlab("Relative Position") +
      ylab("Relative Position") +
      labs(fill = "Residual \nCorrelation") +
      theme(legend.position = c(.1, .8)) +
      annotate("text", x = 0, y = 9, label= paste0(pops[i], " - ", pops[j]))
    print(p)
  }
}
```

### CpG > GpG

```{r, message=FALSE, warning=FALSE}
subtype <- "cpg_GC_CG"
df <- read_subtype_range_resid(subtype)

df_cor <- df %>%
  select(pop, status, rp1, rp2, p1, p2, res) %>%
  pivot_wider(names_from = pop, values_from = res) %>%
  select_if(is.numeric) %>%
  group_by(rp1, rp2) %>%
  nest() %>%
   mutate(
    correlations = map(data, \(x) correlate(x, quiet = T))
  ) %>%
  unnest(correlations) %>%
  select(-data)

pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")

for(i in 1:4){
  for(j in (i+1):5){
    p <- df_cor %>%
      filter(term == pops[i]) %>%
      select(rp1, rp2, !!sym(pops[j])) %>%
      ggplot(aes(x = rp2, y = rp1, fill = !!sym(pops[j]))) +
      geom_tile() +
      scale_fill_distiller(palette = "RdBu", limits = c(-1,1)) +
      theme_bw() +
      xlab("Relative Position") +
      ylab("Relative Position") +
      labs(fill = "Residual \nCorrelation") +
      theme(legend.position = c(.1, .8)) +
      annotate("text", x = 0, y = 9, label= paste0(pops[i], " - ", pops[j]))
    print(p)
  }
}
```



