---
title: "NYGC_control_samples"
author: "Andy Beck"
date: "2023-06-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
base_dir <- "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/controls/"
singlepos_dir <- "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/single_pos/"
```


## Introduction

The purpose of this document is to collect summary statistics regarding the sampled control non-variant sites for each subtype. 

### Data Structure

Inside the directory base directory (`base_dir`) are folders for each 1kGP population (and `ALL`). In each of these, there is a csv file for each subtype (noting here that the files are headerless). The columns of this file are:

1. chromosome
2. singleton position
3. ref
4. ref
5. distance between singleton and control
6. control position
7. 3mer motif centered at control

### Simple functions

```{r}
get_dist_column <- function(subtype, pop, 
                            base_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/controls/"){
  f_name <- paste0(base_dir, pop, "/", subtype, ".csv")
  awk_cmd <- paste0("awk -F, '{print($6)}' ", f_name)
  
  vroom::vroom(pipe(awk_cmd), col_names = "pos", delim = ",")
}

load_sp_res <- function(subtype, pop, suffix = "",
                        base_dir = "/net/snowwhite/home/beckandy/research/1000G_NYGC_LSCI/output/single_pos/"){
  f_name <- paste0(base_dir, pop, "/", subtype, ".csv", suffix)
  read_csv(f_name, show_col_types = FALSE)
}
```

## Distance Distributions

### A > C

```{r, message=FALSE, warning=FALSE}
subtype <- "AT_CG"
df <- get_dist_column(subtype, "ALL")
```

Let's look at a density plot of the distances for this subtype:

```{r}
df %>%
  ggplot(aes(x = pos)) +
  geom_density()
```

### A > G

```{r, message=FALSE, warning=FALSE}
subtype <- "AT_GC"
df <- get_dist_column(subtype, "ALL")
```

Let's look at a density plot of the distances for this subtype:

```{r}
df %>%
  ggplot(aes(x = pos)) +
  geom_density()
```

## Near versus Far Results

### A > C

```{r}
subtype <- "AT_CG"
df_min <- load_sp_res(subtype, "ALL", suffix = ".min")
df_max <- load_sp_res(subtype, "ALL", suffix = ".max")

df_min$re <- df_min$dev / (2* (df_min$singletons + df_min$controls))
df_max$re <- df_max$dev / (2* (df_max$singletons + df_max$controls))
```

Plot results against each other:

```{r}
df_min %>%
  rename(close = re) %>%
  select(offset, close) %>%
  inner_join({
    df_max %>%
      rename(far = re) %>%
      select(offset, far)
  }, by = "offset") %>%
  ggplot(aes(x = close, y = far)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

df_min %>%
  rename(close = re) %>%
  select(offset, close) %>%
  inner_join({
    df_max %>%
      rename(far = re) %>%
      select(offset, far)
  }, by = "offset") %>%
  ggplot(aes(x = log10(close), y = log10(far))) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)
```

### A > G

```{r}
subtype <- "AT_GC"
df_min <- load_sp_res(subtype, "ALL", suffix = ".min")
df_max <- load_sp_res(subtype, "ALL", suffix = ".max")

df_min$re <- df_min$dev / (2* (df_min$singletons + df_min$controls))
df_max$re <- df_max$dev / (2* (df_max$singletons + df_max$controls))
```

```{r}
df_min %>%
  rename(close = re) %>%
  select(offset, close) %>%
  inner_join({
    df_max %>%
      rename(far = re) %>%
      select(offset, far)
  }, by = "offset") %>%
  ggplot(aes(x = close, y = far)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

df_min %>%
  rename(close = re) %>%
  select(offset, close) %>%
  inner_join({
    df_max %>%
      rename(far = re) %>%
      select(offset, far)
  }, by = "offset") %>%
  ggplot(aes(x = log10(close), y = log10(far))) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)
```

### A > T

```{r}
subtype <- "AT_TA"
df_min <- load_sp_res(subtype, "ALL", suffix = ".min")
df_max <- load_sp_res(subtype, "ALL", suffix = ".max")

df_min$re <- df_min$dev / (2* (df_min$singletons + df_min$controls))
df_max$re <- df_max$dev / (2* (df_max$singletons + df_max$controls))
```

```{r}
df_min %>%
  rename(close = re) %>%
  select(offset, close) %>%
  inner_join({
    df_max %>%
      rename(far = re) %>%
      select(offset, far)
  }, by = "offset") %>%
  ggplot(aes(x = close, y = far)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

df_min %>%
  rename(close = re) %>%
  select(offset, close) %>%
  inner_join({
    df_max %>%
      rename(far = re) %>%
      select(offset, far)
  }, by = "offset") %>%
  ggplot(aes(x = log10(close), y = log10(far))) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)
```

### TEST

```{r}
df_min <- load_sp_res("AT_TA", "ALL", suffix = ".min")
df_max <- load_sp_res("AT_GC", "ALL", suffix = ".min")

df_min$re <- df_min$dev / (2* (df_min$singletons + df_min$controls))
df_max$re <- df_max$dev / (2* (df_max$singletons + df_max$controls))
```

```{r}
df_min %>%
  rename(close = re) %>%
  select(offset, close) %>%
  inner_join({
    df_max %>%
      rename(far = re) %>%
      select(offset, far)
  }, by = "offset") %>%
  ggplot(aes(x = close, y = far)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  xlab("AT_TA far") +
  ylab("AT_GC far")

df_min %>%
  rename(close = re) %>%
  select(offset, close) %>%
  inner_join({
    df_max %>%
      rename(far = re) %>%
      select(offset, far)
  }, by = "offset") %>%
  ggplot(aes(x = log10(close), y = log10(far))) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  xlab("AT_TA far") +
  ylab("AT_GC far")
```
